---
title: "Homework 06"
author: "Spencer Pease"
date: "5/18/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(bayesLife)
library(bayesPop)

# Data
data("e0F", package = "wpp2019")
e0_sim_dir <- "./data/e0/sim03092016"
tfr_sim_dir <- "./data/tfr/sim01192018"
pop_sim_dir <- "./data/pop/sim05172020"
mig_file <- "./data/WPP2019_Period_Indicators_Medium.csv"
can_mig_num_file <- "./data/statcan_migration.csv"

e0f_all <- e0F %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    cols = -name,
    names_to = "year",
    values_to = "e0",
    names_pattern = "(.*)-"
  ) %>%
  mutate(year = as.integer(year))

# Control randomness
set.seed(9876)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

e0f_hnd <- e0f_all %>%
  filter(name == "Honduras") %>%
  select(-name) %>%
  mutate(gain = lead(e0) - e0)

knitr::kable(
  e0f_hnd,
  booktabs = TRUE, digits = 3, eval = FALSE,
  col.names = c("Period start", "$e_0$", "Gain"),
  caption = "Honduras female life expectancy at birth and observed gains, 1950-2020"
)

```

### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

dl_gain <- function(l, theta) {

  d1 <- theta[1]
  d2 <- theta[2]
  d3 <- theta[3]
  d4 <- theta[4]

  k <- theta[5]
  z <- theta[6]

  (
    (k       / (1 + exp( (-2*log(9) / d2) * (l - d1         - .5*d2) ))) +
    ((z - k) / (1 + exp( (-2*log(9) / d4) * (l - (d1+d2+d3) - .5*d4) )))
  )

}

ls_err <- function(func, data, obs_vals) {

  function(params) {
    fit_vals <- func(data, params)
    sum((fit_vals - obs_vals)^2)
  }

}

```

In order to model gains in life expectancy at birth, we use the six parameter
double logistic model defined in class.

In order to choose appropriate starting parameters, we look at the
double-logistic curve fit for Honduras, and pick out the appropriate deltas,
$k$, and $z$.

```{r}

e0_sim_mcmc <- get.e0.mcmc(e0_sim_dir)
e0.DLcurve.plot(e0_sim_mcmc, country = "Honduras")

```

```{r}
opt_input <- e0f_hnd %>% slice(-n())
loss_func <- ls_err(dl_gain, opt_input$e0, opt_input$gain)

starting_params <- c(3, 3, 4, 2, 1, 4)
opt_result <- optim(starting_params, loss_func)

opt_params <- opt_result$par
opt_err_var <- opt_result$value

```

From this plot we choose the starting values: $[`r starting_params`]$, where
the fist four items are the deltas, followed by $k$ and $z$.

With these starting values, we use `optim()` to minimize the
least-squares-error, loss of the observed gains vs the fitted gains from the
double logistic gain model, which give us the set of optimized parameters
$[`r round(opt_params, 3)`]$ with an error variance of `r round(opt_err_var, 3)`.


### _Q1.b_

```{r}

# Question 1b -------------------------------------------------------------

e0f_hnd_gains <- e0f_hnd %>%
  rename(obs_gain = gain) %>%
  mutate(fit_gain = dl_gain(e0, opt_params))

```

We can plot use our optimized parameters to get estimates of $e_0$ gain, and
compare them to the observed gains to get a sense of how well the optimization
performed:

```{r}
ggplot(e0f_hnd_gains, aes(x = obs_gain, y = fit_gain)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "cadetblue") +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Fit vs Observed Gains",
    subtitle = "Honduras e(0), 1950-2020",
    x = "Observed Gain (years)",
    y = "Fitted Gain (years)"
  )

```

_I'm seeing a weird problem here where my gain function seems to predict the
same value no matter the input $e_0$, not sure what's going on here._

### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

hnd_e0_2020_mean <- e0f_hnd_gains %>%
  filter(year == 2015) %>%
  select(e0, fit_gain) %>%
  rowSums()

hnd_e0_2020_sd <- sqrt(opt_err_var)

hnd_e0_2020_dist <- qnorm(
  seq(.0005, .9995, .001),
  mean = hnd_e0_2020_mean,
  sd = hnd_e0_2020_sd
)

```

Using our observed life expectancy at birth for 2015-2020 and the gain in the
same period, we can create an analytic predictive distribution of possible life
expectancy at birth for 2020-2025, using the variance from our model:


```{r}
ggplot(enframe(hnd_e0_2020_dist), aes(x = value)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Analytic Predictive Distribution of e(0)",
    subtitle = "Honduras, 2020-2025",
    x = "e(0) (years)",
    y = "Density"
  )

```

```{r}
hnd_e0_2020_tbl <- tibble(
  Mean = hnd_e0_2020_mean,
  Median = median(hnd_e0_2020_dist),
  `2.5% PI` = Mean - 1.96 * hnd_e0_2020_sd,
  `97.5% PI` = Mean + 1.96 * hnd_e0_2020_sd
)

knitr::kable(
  hnd_e0_2020_tbl,
  booktabs = TRUE, digits = 3, eval = FALSE,
  caption = "Summary of the predictive distribution of 2020-2025 Honduras $e_0$"
)

```



## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

```

### _Q2.a_

```{r}

# Question 2a -------------------------------------------------------------

e0_sim_mcmc <- get.e0.mcmc(e0_sim_dir)
e0_sim_pred <- get.e0.prediction(e0_sim_dir)

```

Data from a fully converged simulation from a Bayesian model created with
_BayesLife_ are loaded using the functions `get.e0.mcmc()` and
`get.e0.prediction()`. `get.e0.mcmc()` returns an object containing each MCMC
chain from the simulation, and `get.e0.prediction()` returns an object
containing the summary statistics of the projections created using an input set
of MCMC chains.

### _Q2.b_

**Double logistic curve fits for Algeria and Morocco:**

```{r}

# Question 2b -------------------------------------------------------------

e0.DLcurve.plot(e0_sim_mcmc, "Algeria",
  predictive.distr = TRUE, pi = c(80, 95), ylim = c(-2, 8)
)

e0.DLcurve.plot(e0_sim_mcmc, "Morocco",
  predictive.distr = TRUE, pi = c(80, 95), ylim = c(-2, 8)
)

```

While the overlap of the predictive intervals for both countries suggest that we
can't make strong conclusions about the real difference in female gains in life
expectancy, we do observe that both countries reach their peak 5-year gains
around when life expectancy at birth is 60 years. According to the median,
Algeria has a higher expected 5-year gains on average than Morocco for any given
$e_0$. Also, Morocco also has a wider probability interval for $e_0$ above 70.

### _Q2.c_

```{r}

# Question 2c -------------------------------------------------------------

e0_traj_tbl <- list(Algeria = "Algeria", Morocco = "Morocco") %>%
  purrr::map(~e0.trajectories.table(e0_sim_pred, country = .x)) %>%
  purrr::map_dfr(~as_tibble(.x, rownames = "year"), .id = "country") %>%
  mutate(year = as.integer(year) - 3) %>%
  filter(year < 2015) %>%
  select(country, `Period start` = year, median) %>%
  pivot_wider(names_from = country, values_from = median)

e0_traj_mean_tbl <- e0_traj_tbl %>%
  pivot_longer(-`Period start`, names_to = "Country", values_to = "e0") %>%
  group_by(Country) %>%
  mutate(gain = lead(e0) - e0) %>%
  summarise(`Mean Gain` = mean(gain, na.rm = TRUE)) %>%
  arrange(desc(`Mean Gain`))

```

Looking at the double-logistic plots, it appears that Algeria has been
experiencing faster increases in life expectancy at birth, as reported by the
larger average predicted gains at all values of $e_0$. We can confirm this by
examining the reported life expectancies for each country from 1950 to 2015:

```{r}
knitr::kable(
  e0_traj_tbl,
  booktabs = TRUE, digits = 3, eval = FALSE,
  caption = "Female $e_0$ for Algeria and Morocco, 1950-2015"
)
```

Again, it appears Algeria has a faster increase. We can verify this by
calculating the mean gain over time for each country:

```{r}
knitr::kable(
  e0_traj_mean_tbl,
  booktabs = TRUE, digits = 3, eval = FALSE,
  caption = "Mean gains in female $e_0$ over 1950-2015 for Algeria and Morocco"
)
```

which confirms the previous two observations.

### _Q2.d_

```{r}

# Question 2d -------------------------------------------------------------

e0_dza_traj <- get.e0.trajectories(e0_sim_pred, "Algeria")[-1, ]
e0_mar_traj <- get.e0.trajectories(e0_sim_pred, "Morocco")[-1, ]

dza_gt_mar_each <- rowMeans(e0_dza_traj > e0_mar_traj)
dza_lt_mar_each <- rowMeans(e0_dza_traj < e0_mar_traj)

dza_gt_mar_all <- mean(apply(e0_dza_traj > e0_mar_traj, 2, all))
dza_lt_mar_all <- mean(apply(e0_dza_traj < e0_mar_traj, 2, all))

dza_mar_prob_each_tbl <- tibble(
  period_start = as.integer(names(dza_gt_mar_each)) - 3,
  prob_dza_higher = dza_gt_mar_each,
  prob_dza_lower = dza_lt_mar_each
)

```

Assuming gains in life expectancy are independent between countries, we can find
the probability of Algeria having either a higher or loweraverage life
expectancy at birth than Morocco in each future period by calculating the mean
number of times Algeria has a higher (or lower) $e_0$ than Morocco across all
simulations:

```{r}
knitr::kable(
  dza_mar_prob_each_tbl,
  booktabs = TRUE, digits = 3, eval = FALSE,
  col.names = c("Period start", "Pr(DZA > MAR)", "PR(DZA < MAR)"),
  caption = paste("Probability of Algeria having a higher average female $e_0$",
                  "in each future period")
)
```

Similarly, we can find the probability of $e_0$ either being higher or lower in
Algeria than Morocco in all future periods by finding the fraction of
simulations where each condition is met:

$$
\begin{aligned}
Pr(\text{DZA} > \text{MAR}) &= `r round(dza_gt_mar_all, 3)` \\
Pr(\text{DZA} < \text{MAR}) &= `r round(dza_lt_mar_all, 3)`
\end{aligned}
$$


## _Q3_

```{r}

# Question 3 --------------------------------------------------------------

# Run this once to get pop predictions

# pop_sim_pred <- pop.predict(
#   end.year = 2100, start.year = 1950, present.year = 2020, wpp.year = 2019,
#   output.dir = pop_sim_dir,
#   inputs = list(
#     tfr.sim.dir = tfr_sim_dir,
#     e0F.sim.dir = e0_sim_dir,
#     e0M.sim.dir = "joint_"
#   ),
#   nr.traj = 50,
#   keep.vital.events = FALSE
# )

pop_sim_pred <- get.pop.prediction(pop_sim_dir)

```

### _Q3.a_

```{r}

# Question 3a -------------------------------------------------------------

# Defined using `?pop.expressions`
over65_exp <- "PCAN[14:27]"
support_exp <- "PCAN[5:13] / PCAN[14:27]"

```

Using our converged life expectancy at birth and total fertility rate (from
homework 5) simulations, we can create probabilistic projections of the
following population quantities:

_Note that the "Potential Support Ratio" is defined as
$\frac{\text{people aged 20-64}}{\text{people aged 65 and over}}$_


```{r}
pop.trajectories.plot(
  pop_sim_pred, "Canada",
  sex = "both",
  sum.over.ages = TRUE,
  main = "Canada Total Population"
)

pop.trajectories.plot(
  pop_sim_pred, "Canada",
  sex = "male",
  sum.over.ages = TRUE,
  main = "Canada Total Male Population"
)

pop.trajectories.plot(
  pop_sim_pred, "Canada",
  expression = over65_exp,
  sex = "both",
  sum.over.ages = TRUE,
  main = "Canada Total Population over 65"
)

pop.trajectories.plot(
  pop_sim_pred, "Canada",
  expression = support_exp,
  sex = "both",
  sum.over.ages = TRUE,
  main = "Canada Potential Support Ratio"
)
```

Looking at the projections of potential support ratio, we see continued
declines into the future, though eventually at a slower rate. This indicates
Canada's age structure is likely to continue shifting to older ages.


### _Q3.b_

```{r}

# Question 3b -------------------------------------------------------------

# pop_agg_n_am <- pop.aggregate(pop_sim_pred, regions = 905)
pop_agg_n_am <- get.pop.aggregation(pop_sim_dir)

pop.trajectories.plot(pop_agg_n_am, 905, sum.over.ages = TRUE)

```

We can also aggregate the projected population simulations from multiple
countries to get the probabilistic population projection of a region, like
North America.


## _Q4_

```{r}

# Question 4 --------------------------------------------------------------

can_mig <- readr::read_csv(mig_file) %>%
  filter(Location == "Canada") %>%
  transmute(year = MidPeriod - 3, cnmr = CNMR) %>%
  filter(year < 2020)

```

### _Q4.a_

```{r}

# Question 4a -------------------------------------------------------------

knitr::kable(
  can_mig,
  booktabs = TRUE, digits = 3,
  col.names = c("Period start", "CNMR"),
  caption = "Crude net migration rate (CNMR) for Canada, 1950-2100"
)

```

### _Q4.b_

```{r}

# Question 4b -------------------------------------------------------------

can_mig_ar_model <- arima(can_mig$cnmr, order = c(1, 0, 0), method = "ML")

can_mig_ar_tbl <- tibble(
  Parameter = c("AR(1) Param", "Mean"),
  Value = can_mig_ar_model$coef,
  `S.E` = sqrt(diag(vcov(can_mig_ar_model)))
)

```

After fitting an autoregressive ($AR(1)$) model to the series of crude net
migration rates over time, we get a model with the parameters:

```{r}
knitr::kable(
  can_mig_ar_tbl,
  booktabs = TRUE, digits = 3,
  caption = "AR(1) model parameters for Canada CNMR, 1950-2020"
)
```

### _Q4.c_

```{r}

# Question 4c -------------------------------------------------------------

can_r_2015 <- can_mig %>% filter(year == 2015) %>% pull(cnmr)
can_mu <- can_mig_ar_model$coef[[2]]
can_phi <- can_mig_ar_model$coef[[1]]

can_r_2020_mean <- can_phi * (can_r_2015 - can_mu) + can_mu
can_r_2020_sd <- sqrt(can_mig_ar_model$sigma2)

can_r_2020_dist <- qnorm(
  seq(.0005, .9995, .001),
  mean = can_r_2020_mean,
  sd = can_r_2020_sd
)

```

Using these model parameters, we can find an analytic solution for the
predictive probability distribution of net migration rates in Canada for
2020-2025, which takes the form:

$$
\text{CNMR}_{2020-2025} \sim
N(`r round(can_r_2020_mean, 3)`, `r round(can_r_2020_sd, 3)`)
$$

```{r}
ggplot(enframe(can_r_2020_dist), aes(x = value)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Analytic Predictive Distribution of Crude Net Migration Rate",
    subtitle = "Canada, 2020-2025",
    x = "Crude Net Migration Rate",
    y = "Density"
  )

```


### _Q4.d_

```{r}

# Question 1d -------------------------------------------------------------

can_r_2020_sample <- rnorm(1000, mean = can_r_2020_mean, sd = can_r_2020_sd)

can_r_2020_tbl <-
  tibble(
    Analytic = can_r_2020_dist,
    Sample = can_r_2020_sample
  ) %>%
  pivot_longer(everything(), names_to = "dist", values_to = "value")

```

We can also sample from this distribution to show it does indeed follow what
we predict analytically:

```{r}
ggplot(can_r_2020_tbl, aes(x = value, fill = dist)) +
  geom_histogram(
    data = filter(can_r_2020_tbl, dist == "Sample"),
    aes(y = ..density..),
    binwidth = .5,
    alpha = .5
  ) +
  geom_density(
    data = filter(can_r_2020_tbl, dist == "Analytic"),
    alpha = .2
  ) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predictive Distributions of Crude Net Migration Rate (Analytic and Sample)",
    subtitle = "Canada, 2020-2025",
    x = "Crude Net Migration Rate",
    y = "Density",
    fill = "Type"
  )

```


### _Q4.e_

```{r}

# Question 4e -------------------------------------------------------------

can_mig_num_full <- readr::read_csv(can_mig_num_file) %>%
  select(age_group = `Age group`, sex = `Sex`, type = `Type of migrant`, value = `VALUE`) %>%
  pivot_wider(names_from = "type", values_from = "value") %>%
  mutate(
    net_mig = `Immigrants` - `Emigrants`,
    sex = case_when(
      sex == "Males"      ~ "male",
      sex == "Females"    ~ "female",
      sex == "Both sexes" ~ "both",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-`Immigrants`, -`Emigrants`)

can_mig_prop_both <- can_mig_num_full %>%
  filter(sex == "both") %>%
  select(-sex) %>%
  mutate(
    prop_mig = pmax(net_mig / sum(net_mig), 0),
    mig_rate = can_r_2020_mean * prop_mig
  )

  # extract(age_group, into = "age_group_start", regex = "(\\d+)", convert = TRUE)

```

Using data downloaded from Canada's statistical website (_www.statcan.gc.ca_),
we can get the age-specific net migration number for 2018-2019. From this, we
can then get the proportion of net migration in each age group. _Note that there
are zero values because some age groups had more emigrants than immigrants.
These were coerced to zero, under the assumption that old ages are less likely
to migrate._

```{r}
knitr::kable(
  select(can_mig_prop_both, age_group, mig_rate),
  booktabs = TRUE, digits = 3,
  col.names = c("Age group", "Net migration rate"),
  caption = "Age schedule of net migration rate (per 1000 people)"
)
```


### _Q4.f_

```{r}

# Question 1f -------------------------------------------------------------

# Define ages to get from trajectories
age_map_list <- as.list(1:20)
age_map_list[[21]] <- 21:27 # 100+ age aggregate

can_pop_2020_sim <- age_map_list %>%
  purrr::map(~pop.trajectories(pop_sim_pred, country = "Canada", age = .x)) %>%
  purrr::map_dfr(~as_tibble(.x, rownames = "year"), .id = "age_id") %>%
  filter(year == 2025) %>%
  mutate(age_group = can_mig_prop_both$age_group) %>%
  select(age_group, everything(), -age_id, -year) %>%
  pivot_longer(-age_group, names_to = "sample", values_to = "pop") %>%
  extract(sample, into = "sample", regex = "(\\d+)", convert = TRUE)

can_mig_prop_both_sample <-
  rnorm(50, mean = can_r_2020_mean, sd = can_r_2020_sd) %>%
  purrr::map_dfr(
    ~mutate(can_mig_prop_both, mig_rate = .x * prop_mig),
    .id = "sample"
  ) %>%
  mutate(sample = as.integer(sample)) %>%
  select(sample, age_group, mig_rate) %>%
  left_join(can_pop_2020_sim, by = c("age_group", "sample")) %>%
  mutate(mig_number = mig_rate * pop)


```

```{r fig.asp=1.5, fig.width=7.5}
ggplot(can_mig_prop_both_sample, aes(x = mig_number)) +
  geom_histogram(aes(y = ..density..)) +
  facet_wrap(vars(age_group), ncol = 2) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Projected Age-Specific Net Migration Numbers",
    subtitle = "Canada, 2020-2025",
    x = "Net number of migrants",
    y = "Density"
  )
```

Using our 50 simulated population projections, we can get the probabilistic
projected age-specific net migration numbers by taking a samples from the
$AR(1)$ model and applying the same schedule of age-specific migration for each
simulated trajectory.

### _Q4.g_

```{r}

# Question 4g -------------------------------------------------------------

# use pop.predict with migration included?

```



# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
