---
title: "Homework 01"
author: "Spencer Pease"
date: "4/6/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(ggplot2)

# Make data
age_range <- c(0, 110)
age_data <- tibble(age = seq(age_range[1], age_range[2], .1))

```


## _Q1_

For a given force of mortality function, $\mu(x)$, the cumulative hazard
is defined as the total area under the curve of $\mu(x)$ bounded on the interval
$[0, x]$, or put another way:

$$ \Lambda(x) = \int_0^x \mu(u) du $$
For the force of mortality function $\mu(x) = 0.005 + 0.0005(x - 15)^2$, the
cumulative hazard function is then:

$$
\begin{aligned} \Lambda(x) &= \int_{0}^{x} \mu(u) du \\
&= \int_{0}^{x} \left[0.005 + 0.0005(u - 15)^2 \right]du \\
&= 0.005x + 0.0005 \int_0^x \left(u^2 - 30u + 225\right)du \\
&= 0.005x + 0.0005 \left. \left( \frac{1}{3}u^3 - 15u^2 + 225u \right) \right|_0^x \\
&= 0.005x + 0.0005 \left( \frac{1}{3}x^3 - 15x^2 + 225x \right) \\
&= \frac{1}{2000}\left( \frac{1}{3}x^3 - 15x^2 + 235x \right)
\end{aligned}
$$



For ages `r age_range[1]` to `r age_range[2]`, this cumulative hazard functions
looks like:

```{r question-1}

# Question 1 --------------------------------------------------------------

hazard_fun <- function(x) 0.005 + 0.0005 * (x − 15)^2
cum_hazard_fun <- function(x) .0005 * ((x^3 / 3) - (15 * x^2) + (235 * x))

chf_plot <- ggplot(age_data, aes(x = age, y = cum_hazard_fun(age))) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Cumulative Hazard vs Age",
    x = "Age (years)",
    y = "Cumulative Hazard"
  )

chf_plot

```


## _Q2_

The survival function, $S(x)$, is defined as the exponentiated negative
cumulative hazard function, $e^{-\Lambda(x)}$. Using our calculated cumulative
hazard function, the survival function is then:

$$
S(x) =
\text{exp}\left[
\frac{-1}{2000}\left( \frac{1}{3}x^3 - 15x^2 + 235x \right)
\right]
$$

For ages `r age_range[1]` to `r age_range[2]`, the survival function then looks
like:

```{r question-2}

# Question 2 --------------------------------------------------------------

survival_fun <- function(x) exp(-1 * cum_hazard_fun(x))

survf_plot <- ggplot(age_data, aes(x = age, y = survival_fun(age))) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Survival vs Age",
    x = "Age (years)",
    y = "Survival"
  )

survf_plot

```


## _Q3_

The probability density function of $X$, $f(x)$, is the negative derivative of
the survival function with respect to x, $f(x) = -\frac{dS(x)}{dx}$. Using our
calculated survival function, the probability density function of $X$ is then:

$$
\begin{aligned} f(x) &= \frac{-d}{dx}
\text{exp}\left[\frac{-1}{2000}\left( \frac{1}{3}x^3 - 15x^2 + 235x \right) \right] \\
&= \frac{1}{2000} \frac{d}{dx}\left( \frac{1}{3}x^3 - 15x^2 + 235x \right) S(x) \\
&= \frac{1}{2000} \left( x^2 - 30x + 235 \right) S(x) \\
&= \mu(x)S(x)
\end{aligned}
$$

For ages `r age_range[1]` to `r age_range[2]`, the probability density function
looks like:

```{r question-3}

# Question 3 --------------------------------------------------------------

pdfun_plot <-
  ggplot(age_data, aes(x = age, y = survival_fun(age) * hazard_fun(age))) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Probability density vs Age",
    x = "Age (years)",
    y = "Probability density"
  )

pdfun_plot

```


## _Q4_

```{r question-4}

# Question 4 --------------------------------------------------------------

cohort_size <- 100000

lt_data <- tibble(
  age = 0:50,
  lx = survival_fun(age) * cohort_size,
  dx = lx - lead(lx),
  qx = dx / lx,
  mx =  dx / (lx - .5 * dx)
)

lt_names <- c("age", "$l_x$", "$_{1}d_x$", "$_{1}q_x$", "$_{1}m_x$")

knitr::kable(lt_data,
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  col.names = lt_names,
  caption = "Cohort life table for single year ages 0 to 50"
)

```


## _Q5_

```{r question-5}

# Question 5 --------------------------------------------------------------

e0 <- integrate(survival_fun, lower = 0, upper = Inf)
e0_val <- round(e0$value, 3)

```


Life expectancy at age $x$, $e_x$, is defined as:

$$e_x = \frac{\int_x^{\infty}S(u)du}{S(x)}$$

which simplifies to $\int_0^{\infty}S(u)du$ for life expectancy at birth, $e_0$.
Using numerical integration, the life expectancy at birth for our cohort is
calculated to be **`r e0_val`**.


## _Q6_

```{r question-6}

# Question 6 --------------------------------------------------------------

e10 <- integrate(survival_fun, lower = 10, upper = Inf)
e10_val <- round(e10$value / survival_fun(10), digits = 3)

```

The life expectancy at age 10 ($e_{10}$) for a member of this cohort is
numerically calculated to be **`r e10_val`**.


## _Q7_

The probability that a person aged $x$ dies within the next $n$ years is defined
as:

$$_{n}q_x = \frac{S(x) - S(x + n)}{S(x)}$$

```{r question-7}

# Question 7 --------------------------------------------------------------

nqx <- function(x, n) (survival_fun(x) - survival_fun(x + n)) / survival_fun(x)
q15_15 <- round(nqx(15, 15), 3)

```

The $_{15}q_{15}$ value for this cohort is then **`r q15_15`**.


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
