---
title: "Homework 01"
author: "Ihsan Kahveci"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: yes
    number_sections: true
---

\tableofcontents 
\newpage 


```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')
plot_size=15

rm(gr)
```


```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(tidyverse, quietly = TRUE)
library(wpp2019)

# Make data
age_range <- c(0, 110)
age_data <- tibble(age = seq(age_range[1], age_range[2], .1))

```


# _Q1_

If the instantaneous mortality rate is constant (0.02) --- independent of age, the probability distribution follows an exponential distribution. Let $x$ equals to age at death; hence the relevant probability density function $f(x)$, cumulative distribution function $F(x)$, survival function $S(x)$: 


$$
\begin{aligned}
f(x) = 0.02e^{-0.02x} \\ 
F(x) = 1-e^{0.02x} \\ 
S(x) = e^{-0.02x} 
\end{aligned}
$$ 


```{r q1 }
# Question 1 ---------------------------------------------------------------
pdf = function(x) 0.02*exp(-0.02*x)
cdf = function(x) 1 - exp(-0.02*x)
survf = function(x) exp(-0.02*x)

pdf_plot = 
  age_data %>% 
  mutate(PDF = pdf(age)) %>% 
  ggplot(aes(x=age, y=PDF)) + 
  geom_line() + 
  theme_bw(base_size = plot_size) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Probability Density Function", 
    x = "Age (years)",
    y = "Probability Density"
    )

median_plot = 
  age_data %>%
  mutate(surv = survf(age)) %>%
  ggplot(aes(x=age, y=surv)) + 
  geom_line() + 
  geom_hline(yintercept = 0.5, linetype="dashed") + 
  annotate("text", x=40, y=0.53, label="Median age = 34", color = "blue", size=6) + 
  theme_bw(base_size = plot_size) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Survival Curve", 
    x = "Age (years)",
    y = "Survival Probability"
    )
```



## (a)

Plotting the probability density function:

```{r}
pdf_plot
```


## (b)

Probability that a member of this population is still alive at age 70: $S(70) =$ `r survf(70)`


## (c)
Probability that a member of this population dies before age 6: $F(6) =$ `r cdf(6)` 

## (d)
Life expectancy at birth for a member of this population: $e_{0} =$ `r 1/0.02`  

## (e) 
Life expectancy at age 50 for a member of this population: $e_{50} =$ `r 1/0.02`  

## (f)
Median age at death for this population: $S(50) =$ `r survf(34)`

```{r}
median_plot
```


# _Q2_

```{r q2}
# Question 2 ---------------------------------------------------------------
hazard_fun = function(x) (0.0168*x^2 - 0.668*x + 8)/1000

hazard_plot = 
  age_data %>%
  mutate(mortality = hazard_fun(age)) %>%
  ggplot(aes(x=age, y=mortality)) + 
  geom_line() + 
  theme_bw(base_size = plot_size) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Mortality over Age", 
    x = "Age (years)",
    y = "Hazard rate"
    )

cum_hazard_fun <- function(x) x*(14*x^2 - 835*x + 20000)/250000

chf_plot <- 
  ggplot(age_data, aes(x = age, y = cum_hazard_fun(age))) +
  geom_line() +
  theme_bw(base_size = plot_size) +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Cumulative Hazard vs Age",
    x = "Age (years)",
    y = "Cumulative Hazard"
  )

survival_fun <- function(x) exp(-1 * cum_hazard_fun(x))

surv_plot = 
  age_data %>%
  mutate(surv = survival_fun(age)) %>%
  ggplot(aes(x=age, y=surv)) + 
  geom_line() + 
  theme_bw(base_size = plot_size) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Survival Curve", 
    x = "Age (years)",
    y = "Survival Probability"
    )

pdf_fun = function(x) survival_fun(x) * hazard_fun(x)

pdfun_plot <-
  ggplot(age_data, aes(x = age, y = pdf_fun(age))) +
  geom_line() +
  theme_bw(base_size = plot_size) +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Probability density vs Age",
    x = "Age (years)",
    y = "Probability density"
  )


e0 <- integrate(survival_fun, lower = 0, upper = Inf)
e0_val <- round(e0$value, 3)

e10 <- integrate(survival_fun, lower = 10, upper = Inf)
e10_val <- round(e10$value / survival_fun(10), digits = 3)

nqx <- function(x, n) (survival_fun(x) - survival_fun(x + n)) / survival_fun(x) 
q45_15 <- round(nqx(15, 45), 3)
```

## (a)

For ages `r age_range[1]` to `r age_range[2]`, this mortality rate plot
looks like:

```{r}
hazard_plot
```


## (b)

For a given instantaneous mortality function $\mu(x) = (0.0168x^2 − 0.668x + 8)/1000$, defined as the total area under the curve of $\mu(x)$ bounded on the interval $[0, x]$, or put another way:

$$
\begin{aligned} \Lambda(x) &= \int_{0}^{x} \mu(u) du \\
&= \int_{0}^{x} \left[(0.0168x^2 − 0.668x + 8)/1000 \right]dx \\ 
&= \int_{0}^{x} \dfrac{21x^2-835x+10000}{1250000}dx\\
&= \dfrac{21}{125000}\int_{0}^{x}x^2dx - \frac{167}{250000}\int_{0}^{x}xdx + \frac{1}{125}\int_{0}^x1dx \\
&= \dfrac{x\cdot\left(14x^2-835x+20000\right)}{250000} + C
\end{aligned}
$$
For ages `r age_range[1]` to `r age_range[2]`, this cumulative hazard functions
looks like:

```{r}
chf_plot
```


The survival function, $S(x)$, is defined as the exponentiated negative
cumulative hazard function, $e^{-\Lambda(x)}$. Using our calculated cumulative
hazard function, the survival function is then:

$$
S(x) =
\text{exp}\left[
\dfrac{-x\cdot\left(14x^2-835x+20000\right)}{250000}
\right]
$$

## (c)
For ages `r age_range[1]` to `r age_range[2]`, the survival function then looks
like:  

```{r}
surv_plot
```

## (d)

The probability density function of $X$, $f(x)$, is the negative derivative of
the survival function with respect to x, $f(x) = -\frac{dS(x)}{dx}$. Using our
calculated survival function, the probability density function of $X$ is then:

$$
\begin{aligned} f(x) &= \frac{-d}{dx} = \mu(x)S(x)
\end{aligned}
$$

For ages `r age_range[1]` to `r age_range[2]`, the probability density function
looks like:

```{r}
pdfun_plot

```


## (e)


Life expectancy at age $x$, $e_x$, is defined as:

$$e_x = \frac{\int_x^{\infty}S(u)du}{S(x)}$$

which simplifies to $\int_0^{\infty}S(u)du$ for life expectancy at birth, $e_0$.
Using numerical integration, the life expectancy at birth for our cohort is
calculated to be **`r e0_val`**.


## (f)

The life expectancy at age 10 ($e_{10}$) for a member of this cohort is
numerically calculated to be **`r e10_val`**.


## (g)

The probability that a person aged $x$ dies within the next $n$ years is defined
as:

$$_{n}q_x = \frac{S(x) - S(x + n)}{S(x)}$$
The $_{45}q_{15}$ value for this cohort is then **`r q45_15`**.


# _Q3_

## (a) & (b)

$$ _nd_x = l_x - l_{x+n}$$

```{r}
# Question 3 ---------------------------------------------------------------
## download estimates of males
data("popM")

peru = 
  popM %>%
  filter(name == "Peru") %>%
  filter(age == c("70-74", "75-79", "80-84", "85-89", "90-94", "95-99", "100+")) %>%
  select(age, "2015", "2020") %>%
  mutate(ndx = `2015` - lead(`2020`)) %>%
  filter(age != "100+") 




peru %>%
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "UN Prospects for Peru")

```



# _Q4_

## (a) 

$$_nM_x = \frac{_nd_x}{_nL_x}$$
```{r}
# Question 4 ---------------------------------------------------------------
peru_LT =
  peru %>%
  mutate(nLx = (5*(`2015`+`2020`)/2)) %>%
  mutate(nMx = ndx/nLx) 
  
peru_LT %>% 
 knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "UN Life Table for Peru")
  
```

## (b)

$$ 
\begin{aligned}
\text{Solution 1:} \\ 
_nq_x \approx 1 - e^{-_nM_x} \\
\text{   }\\
\text{Solution 2:} \\
_nq_x \approx \frac{n_nM_x}{1 + n_nM_x/2}
\end{aligned}
$$
\newpage 

```{r}
## first method
peru_LT %>%
 mutate(nqx = 1-exp(-5*nMx)) %>%
 knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "First Method", 
              col.names=c("age", "2015", "2020", "$_nd_x$", "$_nL_x$", "$_nM_x$", "$_nq_x$"))


## second method
peru_LT %>%
  mutate(nqx = (5*nMx)/(1+5*nMx/2)) %>%
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "Second Method",
               col.names=c("age", "2015", "2020", "$_nd_x$", "$_nL_x$", "$_nM_x$", "$_nq_x$"))
```
## (c)

$$_nm_x = \frac{_nd_x}{_nL_x} = \frac{_nd_x}{nl_x − 1/2*n_nd_x}$$
```{r}
peru_LT %>% 
  mutate(nLx = 5*`2015`-(1/2)*5*ndx) %>%
  mutate(nmx = ndx/nLx) %>%
  select(-nMx) %>%
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "Life Table $_nm_x$", 
               col.names = c("age", "2015", "2020", "$_nd_x$", "$_nL_x'$", "$_nm_x$"))

```

## (d)

```{r}
## comparison
peru_LT %>%
  mutate(nqx_1 = 1-exp(-5*nMx)) %>%
  mutate(nqx_2 = (5*nMx)/(1+5*nMx/2)) %>%
  select(age, nqx_1, nqx_2) %>%
  knitr::kable(digits = 3, booktabs = TRUE, escape = FALSE,
               caption = "Comparison", col.names = c("age", "$_nq_x^1$", "$_nq_x^2$"))
```

As seen from the table above, the estimates from two methods are quite similiar. The difference occurs in later ages because $_nM_x$ becomes larger. I think second method is better because it doesn't assume that mortality rate is constant within the age interval. Instead, it is modeled as a linear function of age. As we move from 5-year age groups to single age groups, this became more important because even two-year age difference matters in later ages. 


## (e)

```{r}
peru_LT %>%
  mutate(nLx2 = 5*`2015`-(1/2)*5*ndx) %>%
  mutate(nmx = ndx/nLx2) %>%  
  select(age, `2015`, ndx, nLx, nLx2, nMx, nmx) %>%
  knitr::kable(digits = 3, booktabs = TRUE, escape = FALSE,
               caption = "Comparison: $_nM_x$ and $_nm_x$", 
               col.names = c("age", "2015", "$_nd_x$", "$_nL_x$", "$_nL_x'$" , "$_nM_x$", "$_nm_x$"))

```

$_nM_x$ uses person-years $(_nL_x)$ assuming mortality rate is constant within 5-year age groups. "$_nm_x$" uses person-years $(_nL_x')$ assuming mortality rate is a linear function of age. As a result, former estimates more person-years because it doesn't take into account the deaths occur within the early years of the age group which happens more in older ages. As expected, $_nM_x$ underestimates mortality rate in older ages comparing to $_nm_x$

\newpage
# _Q5_
 
 
```{r}
# Question 5 ---------------------------------------------------------------
## Life Table using first method
LT_1 = peru_LT %>%
  mutate(nqx = 1-exp(-5*nMx)) %>%
  select(age, nqx) %>%
  mutate(n = 5, npx = 1-nqx) %>%
  bind_rows(a=list(age="100+", n=5, nqx=1, npx=0))

npx <- LT_1$npx

age <- seq(70,100,5)
lx <- 10000

for (a in age[-length(age)]) {
  l <- lx[which(age == a)] * npx[which(age == a)]
  lx <- c(lx, l)
}

LT_1 <- LT_1 %>%
  mutate(
    x = age,
    lx = lx,
    ndx = lx*nqx,
    Lx = n*lx - n*ndx/2,
    Tx = rev(Lx) %>% coalesce(0) %>% cumsum() %>% rev(),
    ex = Tx / lx
)

## Life Table using second method

LT_2 = peru_LT %>%
  mutate(nqx = (5*nMx)/(1+5*nMx/2)) %>%
  select(age, nqx) %>%
  mutate(n = 5, npx = 1-nqx) %>%
  bind_rows(a=list(age="100+", n=5, nqx=1, npx=0))

npx <- LT_2$npx

age <- seq(70,100,5)
lx <- 10000

for (a in age[-length(age)]) {
  l <- lx[which(age == a)] * npx[which(age == a)]
  lx <- c(lx, l)
}

LT_2 <- LT_2 %>%
  mutate(
    x = age,
    lx = lx,
    ndx = lx*nqx,
    Lx = n*lx - n*ndx/2,
    Tx = rev(Lx) %>% coalesce(0) %>% cumsum() %>% rev(),
    ex = Tx / lx
)
```

## (a)

```{r}
LT_1 %>% 
  select(-n, -x) %>% 
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "Life Table - First Method", 
               col.names=c("age", "$_nq_x$", "$_np_x$", "$_nd_x$", "$l_x$", "$L_x$", "$T_x$", "$e_x$"))


LT_2 %>%    
  select(-n, -x) %>% 
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "Life Table - Second Method",
               col.names=c("age", "$_nq_x$", "$_np_x$", "$_nd_x$", "$l_x$", "$L_x$", "$T_x$", "$e_x$"))

```
## (b)

```{r}
LT_1 %>%
  select(age, ex_1 = ex) %>%
  cbind(ex_2 = LT_2$ex) %>%
  knitr::kable(digits = 2, booktabs = TRUE, escape = FALSE,
               caption = "Life Expectancy Comparison", col.names = c("age", "$e_x^1$", "$e_x^2$" ))
  
```

The first approximation method generated slightly higher life expectancies for all ages, except 100+. This happens because the first method estimates lower probability of death ($_nq_x$).

\newpage 

# Code Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
