---
title: "Homework 05"
author: "Spencer Pease"
date: "5/11/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(R2jags)
library(bayesTFR)

# Data
data("egsingle", package = "mlmRev")
data("tfr", package = "wpp2019")
tfr_sim_dir <- "./data/sim01192018"

tfr_all <- tfr %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    -name,
    values_to = "tfr",
    names_pattern = "^(.*)-",
    names_to = "year",
    names_ptypes = list(year = numeric())
  )

# Control randomness
set.seed(9876)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

edu_data <- egsingle %>%
  filter(year == .5) %>%
  select(childid, schoolid, math) %>%
  pivot_wider(names_from = "schoolid", values_from = "math") %>%
  tibble::column_to_rownames("childid") %>%
  as.matrix()

```

### _Q1.a_

For a student $i$ in school $j$, our Bayesian random effects one-way analysis
of variance model is written as:

$$
\begin{aligned}
y_i &= \alpha_{j[i]} + \epsilon_i, \\
\epsilon_i &\stackrel{iid}{\sim} N(0, \sigma_y^2), \\
\alpha_j   &\stackrel{iid}{\sim} N(\mu_\alpha, \sigma_\alpha^2)
\end{aligned}
$$

where the standard deviation of error in estimating individual student
performance ($\sigma_y$), the mean performance across all schools
($\mu_\alpha$), and the standard deviation in performance across all schools
($\sigma_\alpha$) are the unknown parameters to be estimated.

### _Q1.b_

For this scenario, we set the prior distributions to:

$$
\begin{aligned}
\mu_\alpha    &\sim N(0, 0.0001) \\
\sigma_\alpha &\sim \text{Gamma}(1, 0.0001) \\
\sigma_y      &\sim \text{Gamma}(1, 0.0001)
\end{aligned}
$$

### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

n_students <- dim(edu_data)[1]
n_schools <- dim(edu_data)[2]

edu_model <- jags.model(
  "./edu_jags_model.txt",
  data = list(Y = edu_data, n_students = n_students, n_schools = n_schools),
  n.chains = 3
)

edu_sample_priors <- coda.samples(
  model = edu_model,
  variable.names = c("mu_alpha", "sig_y", "sig_alpha"),
  n.iter = 3000,
  thin = 10
)

edu_sample_schools <- coda.samples(
  model = edu_model,
  variable.names = "alpha_j",
  n.iter = 3000,
  thin = 10
)

```

This model was fit using JAGS.

### _Q1.d_

```{r fig.asp=1.5, fig.width=7.5}

# Question 1d -------------------------------------------------------------

plot(edu_sample_priors)

```

Running the model for 3000 iterations is enough to achieve convergence.

### _Q1.e_

```{r}

# Question 1e -------------------------------------------------------------

edu_quantiles_tbl <- summary(edu_sample_priors)[["quantiles"]] %>%
  as_tibble() %>%
  mutate(Parameter = c("mu", "sigma alpha", "sigma y")) %>%
  select(Parameter, everything())


knitr::kable(
  edu_quantiles_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Summary of posterior distribution of prior parameters"
)

```

We can get the summary of our posterior distributions, and look at the density
plots of the distributions in question _(1.d)_.

## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

hnd_tfr <- tfr_all %>% filter(name == "Honduras") %>% select(-name)

```


### _Q2.a_

```{r}

# Question 2a -------------------------------------------------------------

# run.tfr.mcmc() or nls()?

```

### _Q2.b_
### _Q2.c_

## _Q3_

```{r}

# Question 3 --------------------------------------------------------------

nld_tfr <- tfr_all %>% filter(name == "Netherlands") %>% select(-name)

```

### _Q3.a_

```{r}

# Question 3a -------------------------------------------------------------

nld_phase3_year <- nld_tfr %>%
  arrange(year) %>%
  filter(tfr < 2) %>%
  mutate(
    year_diff = lead(year) - year,
    period_5 = year_diff == 5 & lag(year_diff) == 5,
    two_increases = tfr > lag(tfr, 1) & tfr < lead(tfr, 1)
  ) %>%
  filter(period_5 & two_increases) %>%
  slice(1) %>%
  pull(year)

```

```{r}
knitr::kable(
  nld_tfr,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Period Start", "TFR"),
  caption = "Total fertility rates, Netherlands, 1950-2020"
)
```

The start of Phase III of the fertility model is defined by two consecutive
five-year increases of TFR while staying below a TFR of 2. Looking at TFR data
for the Netherlands, we see that Phase III starts with the period beginning
in **`r nld_phase3_year`**.

### _Q3.b_

```{r}

# Question 3b -------------------------------------------------------------

nld_model <- nld_tfr %>%
  filter(year >= nld_phase3_year) %>%
  pull(tfr) %>%
  ar(aic = FALSE, order.max = 1, method = "mle")

nld_model_results <- tibble(
  Mean = nld_model$x.mean,
  `AR param.` = nld_model$ar,
  `Error var.` = nld_model$var.pred
)

```

We now fit an order $1$ autoregressive model to the subset of Netherlands TFR
data in Phase III, and extract some model parameters below.
_Note that the AR(1) model was fit using the "mle" method._

```{r}
knitr::kable(
  nld_model_results,
  booktabs = TRUE,
  digits = 3,
  caption = "Netherlands Phase III AR(1) model parameters"
)
```

### _Q3.c_

```{r}

# Question 3c -------------------------------------------------------------

nld_ar_rho <- nld_model$ar
rep_tfr <- 2.1
nld_tfr_2015 <- nld_tfr %>% filter(year == 2015) %>% pull(tfr)

nld_pred_mean <- rep_tfr + nld_ar_rho * (nld_tfr_2015 - rep_tfr)
nld_pred_sd <- sqrt(nld_model$var.pred)

nld_pred_dist <- qnorm(seq(.001, .999, .001), mean = nld_pred_mean, sd = nld_pred_sd)

nld_pred_tbl <- tibble(
  Mean = nld_pred_mean,
  Median = median(nld_pred_dist),
  `2.5% PI` = Mean - 1.96 * nld_pred_sd,
  `97.5% PI` = Mean + 1.96 * nld_pred_sd
)

```

A post-transition (Phase III) model of TFR change, as proposed in
_Lee RD, Tuljapurkar S (1994), Stochastic population forecasts for the
United States: beyond high, medium, and low_, is defined as:

$$
f_{c,t+1} \sim N(\mu + \rho(f_{c,t} - \mu), s^2)
$$

where $f_{c,t}$ is the TFR of country $c$ in the five-year period starting at
$t$, $\mu$ is the approximate replacement-level fertility $2.1$, $\rho$ is the
autoregressive parameter, and $s$ is the standard deviation of the random
errors.

Plugging in these values, we analytically find the distribution of Netherlands
TFR for 2020-2025 to be:

$$
f_{c,2020} \sim N(`r round(nld_pred_mean, 3)`, `r round(nld_pred_sd, 3)`^2)
$$

```{r}
knitr::kable(
  nld_pred_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Predictive distribution summary of Netherlands TFR, 2020-2025"
)

```


```{r}
ggplot(tibble(nld_pred_dist), aes(x = nld_pred_dist)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predictive Distribution of TFR",
    subtitle = "Netherlands, 2020-2025",
    x = "Predicted TFR",
    y = "Density"
  )

```


## _Q4_

```{r}

# Question 4 --------------------------------------------------------------

tfr_dza_mar <- tfr_all %>%
  filter(name %in% c("Algeria", "Morocco")) %>%
  pivot_wider(names_from = name, values_from = "tfr") %>%
  rename(`Period Start` = year)

```

### _Q4.a_

```{r}

# Question 4.a ------------------------------------------------------------

tfr_phase2_mcmc  <- get.tfr.mcmc(tfr_sim_dir)
tfr_phase3_mcmc <- get.tfr3.mcmc(tfr_sim_dir)
tfr_pred  <- get.tfr.prediction(tfr_sim_dir)

```

The fully converged simulation is loaded using the `README` file contained with
the data.

### _Q4.b_

```{r}

# Question 4.b ------------------------------------------------------------

DLcurve.plot(
  tfr_phase2_mcmc,
  country = "Algeria",
  nr.curves = 50,
  pi = 95
)

DLcurve.plot(
  tfr_phase2_mcmc,
  country = "Morocco",
  nr.curves = 50,
  pi = 95
)

```

From the above graphs comparing the Phase II double logistic models for
Algeria and Morocco, we see that the TFR decrements maintains higher values
across TFR in Algeria for both the median and 95% PI, suggesting that
fertility is declining faster in Algeria than Morocco. We can also observe
this trend by looking directly at the TFR for both countries over time:

```{r}
knitr::kable(
  tfr_dza_mar,
  booktabs = TRUE,
  digits = 3,
  caption = "TFR over time"
)
```


### _Q4.c_

```{r}

# Question 4c -------------------------------------------------------------

tfr_traj_dza <- get.tfr.trajectories(tfr_pred, country = "Algeria")[-(1:2), ]
tfr_traj_mar <- get.tfr.trajectories(tfr_pred, country = "Morocco")[-(1:2), ]

prob_tfr_dza_higher <- rowMeans(tfr_traj_dza > tfr_traj_mar)

prob_tfr_dza_higher_tbl <- tibble(
  period_start = as.integer(names(prob_tfr_dza_higher)) - 3,
  prob_dza_higher = prob_tfr_dza_higher
)

prob_tfr_dza_higher_all <-
  sum(apply(tfr_traj_dza > tfr_traj_mar, 2, all)) / ncol(tfr_traj_dza)

```

By getting the TFR trajectories for Algeria and Morocco, we can find the
posterior predictive probability for many conditions.

First, we can determine the probability of Algeria having a higher TFR than
Morocco in each five-year period from 2020 through 2095 by finding the mean
number of times Algeria has a higher TFR than Morocco across all simulations:

```{r}
knitr::kable(
  prob_tfr_dza_higher_tbl,
  booktabs = "TRUE",
  digits = 3,
  col.names = c("Period Start", "Pr(DZA > MAR)")
)
```

We can also find the probability that the TFR of Algeria will be higher than that
of Morocco in all five-year periods from 2020 through 2095 by finding the
mean number of simulations where Algeria has a higher TFR than Morocco for all
periods. This value is calculated to be
**`r round(prob_tfr_dza_higher_all, 3)`**.


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
