---
title: "Homework 04"
author: "Ihsan Kahveci"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: yes
    number_sections: true
---

\tableofcontents 
\newpage 

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```


```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------
# Load libraries
library(tidyverse)
library(forecast)

# Data
data("tfr", package = "wpp2019")

tfr_all <- tfr %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    -name,
    values_to = "tfr",
    names_pattern = "^(.*)-",
    names_to = "year") %>%
  mutate(year = as.numeric(year))

# Control randomness
set.seed(57)

```



## _Q3_ 

__Obtain the values of TFR for the Netherlands for 1950-2020 from the 2019 World Population Prospects.__

```{r}

# Question 3 --------------------------------------------------------------
nld_tfr <- tfr_all %>% filter(name == "Netherlands") %>% select(-name)

knitr::kable(
  nld_tfr,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Period Start", "TFR"),
  caption = "Total fertility rates, Netherlands, 1950-2020")

```

### _Q3.a_

__Identify the period in which the Netherlands entered Phase III of the fertility model.__

```{r}
# Question 3a -------------------------------------------------------------

nld_phase3_year <- nld_tfr %>%
  arrange(year) %>%
  filter(tfr < 2) %>%
  mutate(
    year_diff = lead(year) - year,
    period_5 = year_diff == 5 & lag(year_diff) == 5,
    two_increases = tfr > lag(tfr, 1) & tfr < lead(tfr, 1)
  ) %>%
  filter(period_5 & two_increases) %>%
  slice(1) %>%
  pull(year)

```

> "The countries that have entered Phase III are defined as the countries in which two subsequent five-year increases below a TFR of 2 children have been observed." (Alkema et al. 2011)

Looking at TFR chart for the Netherlands, we see that Phase III starts with the period beginning
in **`r nld_phase3_year`**.


```{r}
nld_tfr %>% 
  mutate(phase = case_when(
    year < 1960 ~ "Phase I",
    year >= 1985 ~ "Phase III", 
    TRUE ~ "Phase II")) %>%
  ggplot(aes(year, tfr, color = phase, group = 1)) + 
  geom_line(size = 1, alpha = 0.5) + 
  geom_point(size = 2) + 
  geom_hline(yintercept = 2, linetype = "dashed", size = 1) +
  labs(title = "Total fertility rates, Netherlands, 1950-2020",
       x = " ", y = "TFR") + 
  scale_x_continuous(n.breaks=8) +
  scale_y_continuous(n.breaks=6, limits = c(0, 4)) +
  theme_bw(base_size = 15) +
  theme(legend.title = element_blank(), 
        legend.position = c(0.81, 0.8), 
        legend.background = element_blank(),
        legend.text = element_text(size=10))

```


\newpage 

### _Q3.b_

**Fit a (non-Bayesian) AR(1) model to the Phase III data, estimating the long-term mean, autoregressive parameter, and error variance.**




```{r}
# Question 3b -------------------------------------------------------------
nld_ts <- nld_tfr %>%
  filter(year >= nld_phase3_year) %>%
  pull(tfr) %>%
  ts()

nld_model = arima(nld_ts, order=c(1,0,0))
```


I fit an order $1$ autoregressive model to the subset of Netherlands TFR
data in Phase III, and extract some model parameters below.
_Note that the AR(1) model was fit using the "mle" method._

```{r, echo=FALSE}
modelsummary::msummary(nld_model, 
                       estimate = "{estimate} [{conf.low}, {conf.high}]", 
                       statistic = NULL, fmt = 2, output = "kableExtra",
                       caption = "Netherlands Phase III AR(1) model parameters")  %>%
  kableExtra::kable_styling(latex_options = "hold_position")

```



\newpage 

### _Q3.c_

**Find the predictive distribution of Netherlands TFR for 2020-2025 conditional on this model, analytically or by simulation. Plot the distribution and give its median and a 95% prediction interval.**

$$ (F_{t+1} - \mu) = \rho*(F_{t} - \mu) + \epsilon $$
$$ \sigma^2 =\frac{ \sum^{N}_{i=1} (\hat{y_i} - y_i)^2 }{N} $$

```{r}
nld_tfr_last = nld_tfr$tfr[length(nld_tfr$tfr)]
nld_sd = sqrt(nld_model$sigma2)
nld_mean = nld_model$coef["intercept"]
nld_ar1 = nld_model$coef["ar1"]
nld_mean_pred = nld_ar1 * (nld_tfr_last - nld_mean) + nld_mean
nld_pred_dist <- qnorm(seq(.001, .999, .001), mean = nld_mean_pred, sd = nld_sd)

nld_pred_tbl <- tibble(
  Mean = mean(nld_pred_dist),
  Median = median(nld_pred_dist),
  `2.5% PI` = Mean - 1.96 * nld_sd,
  `97.5% PI` = Mean + 1.96 * nld_sd)
```

- $\sigma^2 = `r nld_model$sigma2`$  
- $F_{2020} = `r nld_mean_pred`$ 

```{r, echo=TRUE}
knitr::kable(
  nld_pred_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Predictive distribution of Netherlands TFR, 2020-2025") %>%
  kableExtra::kable_styling(latex_options = "hold_position")

```
\newpage

```{r, echo=FALSE}
ggplot(tibble(nld_pred_dist), aes(x = nld_pred_dist)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predictive Distribution of TFR",
    subtitle = "Netherlands, 2020-2025",
    x = "Predicted TFR",
    y = "Density"
  )

```




