---
title: "Homework 04"
author: "Ihsan Kahveci"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: yes
    number_sections: true
---

\tableofcontents 
\newpage 

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------
# Load libraries
library(tidyverse)
library(rethinking)
library(bayesTFR)
options(mc.cores = parallel::detectCores())

# Data
data("egsingle", package = "mlmRev")
data("tfr", package = "wpp2019")
tfr_sim_dir <- "TFR/sim01192018"

tfr_all <- tfr %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    -name,
    values_to = "tfr",
    names_pattern = "^(.*)-",
    names_to = "year") %>%
  mutate(year = as.numeric(year))

# Control randomness
set.seed(57)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

edu_data <- egsingle %>%
  filter(year == .5) %>%
  select(childid, schoolid, math) 

```



### _Q1.a_

For a student $i$ in school $j$, our Bayesian random effects one-way analysis
of variance model is written as:

$$
\begin{aligned}
y_i &= \alpha_{j[i]} + \epsilon_i, \\
\epsilon_i &\stackrel{iid}{\sim} N(0, \sigma_y^2), \\
\alpha_j   &\stackrel{iid}{\sim} N(\mu_\alpha, \sigma_\alpha^2)
\end{aligned}
$$

where the standard deviation of error in estimating individual student
performance ($\sigma_y$), the mean performance across all schools
($\mu_\alpha$), and the standard deviation in performance across all schools
($\sigma_\alpha$) are the unknown parameters to be estimated.

### _Q1.b_

For this scenario, I used uninformative priors because there is no a priori information about the schools or students:

$$
\begin{aligned}
\mu_\alpha    &\sim N(0, 0.0001) \\
\sigma_\alpha &\sim \text{Uniform}(0, 1) \\
\sigma_y      &\sim \text{Uniform}(0, 1)
\end{aligned}
$$

### _Q1.c_

```{r, warning=FALSE, message=FALSE}
# Question 1c -------------------------------------------------------------

dat <- list(
  math = edu_data$math,
  student = edu_data$childid,
  school = edu_data$schoolid)

m1 = ulam(
  alist(
    y ~ dnorm(alpha_j, sig_y),
    alpha_j ~ dnorm(mu_alpha, sig_alpha),
    # Priors
    mu_alpha ~ dnorm(0, 0.0001),
    sig_alpha ~ uniform(0, 1),
    sig_y ~ uniform(0, 1)
    ), data=dat, chains=4)
```


### _Q1.d_

```{r}
traceplot(m1)
trankplot(m1)
```
The model is converged in 1000 iterations.

### _Q1.e_

```{r}
out = rethinking::precis(m1, depth = 2)  
names = rownames(out)

summary = bind_cols(vars = names, as_tibble(out))

knitr::kable(
  summary,
  booktabs = TRUE,
  digits = 3,
  caption = "Predictive distribution sumeduy of Netherlands TFR, 2020-2025"
)
```


```{r}
pairs(m1)
```

This model was fit using STAN which uses Hamiltonian Monte Carlo algorithm.


## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

peru_tfr <- tfr_all %>% filter(name == "Peru") %>% select(-name)

```


### _Q2.a_ & _b_

Using Adrian's demo example:

```{r}
# Question 2a -------------------------------------------------------------
tfr  <- peru_tfr$tfr

dl <- function (x, d, a1, a2, a3, a4) {
  d / (1+exp (-(x-a2)/a1)) - d / (1+exp(-(x-a4)/a3)) }

par (mfrow=c(1,2), lwd=2)

d <- 1
a4 <- 6
a3 <- 0.5
a2 <- 2
a1 <- 0.5
g <- dl(tfr,d,a1,a2,a3,a4)

plot (tfr, g, type="line", xlim=rev(c(0,9)), col="blue")
plot (tfr, g, type="p", xlim=rev(c(0,9)), col="green")

```

The plots shows a great fit. As expected, the third phase region of the plot is blank because Peru has not reached Phase 3 yet.

### _Q2.c_


## _Q3_

```{r}

# Question 3 --------------------------------------------------------------

nld_tfr <- tfr_all %>% filter(name == "Netherlands") %>% select(-name)

```

### _Q3.a_

```{r}

# Question 3a -------------------------------------------------------------

nld_phase3_year <- nld_tfr %>%
  arrange(year) %>%
  filter(tfr < 2) %>%
  mutate(
    year_diff = lead(year) - year,
    period_5 = year_diff == 5 & lag(year_diff) == 5,
    two_increases = tfr > lag(tfr, 1) & tfr < lead(tfr, 1)
  ) %>%
  filter(period_5 & two_increases) %>%
  slice(1) %>%
  pull(year)

```

```{r}
knitr::kable(
  nld_tfr,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Period Start", "TFR"),
  caption = "Total fertility rates, Netherlands, 1950-2020"
)
```

The start of Phase III of the fertility model is defined by two consecutive
five-year increases of TFR while staying below a TFR of 2. Looking at TFR data
for the Netherlands, we see that Phase III starts with the period beginning
in **`r nld_phase3_year`**.

### _Q3.b_

```{r}

# Question 3b -------------------------------------------------------------

nld_ts <- nld_tfr %>%
  filter(year >= nld_phase3_year) %>%
  pull(tfr) %>%
  ts(start = 1985, end=2015, frequency = 1)

nld_model = arima(nld_ts, order=c(1,0,0))


```




I fit an order $1$ autoregressive model to the subset of Netherlands TFR
data in Phase III, and extract some model parameters below.
_Note that the AR(1) model was fit using the "mle" method._

```{r}
knitr::kable(
  broom::tidy(nld_model),
  booktabs = TRUE,
  digits = 3,
  caption = "Netherlands Phase III AR(1) model parameters"
)
```

### _Q3.c_

```{r}
nld_sd = sqrt(nld_model$sigma2)
nld_mean = nld_model$coef[2]
nld_pred_dist <- qnorm(seq(.001, .999, .001), mean = nld_mean, sd = nld_sd)

nld_pred_tbl <- tibble(
  Mean = mean(nld_pred_dist),
  Median = median(nld_pred_dist),
  `2.5% PI` = Mean - 1.96 * nld_sd,
  `97.5% PI` = Mean + 1.96 * nld_sd)
```


```{r}
knitr::kable(
  nld_pred_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Predictive distribution sumeduy of Netherlands TFR, 2020-2025"
)

```


```{r}
ggplot(tibble(nld_pred_dist), aes(x = nld_pred_dist)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predictive Distribution of TFR",
    subtitle = "Netherlands, 2020-2025",
    x = "Predicted TFR",
    y = "Density"
  )

```


## _Q4_

```{r}

# Question 4 --------------------------------------------------------------

tfr_peru_ecu <- tfr_all %>%
  filter(name %in% c("Peru", "Ecuador")) %>%
  pivot_wider(names_from = name, values_from = "tfr") %>%
  rename(`Period Start` = year)

```

### _Q4.a_

```{r}

# Question 4.a ------------------------------------------------------------

tfr_phase2_mcmc  <- get.tfr.mcmc(tfr_sim_dir)
tfr_phase3_mcmc <- get.tfr3.mcmc(tfr_sim_dir)
tfr_pred  <- get.tfr.prediction(tfr_sim_dir)

```

The fully converged simulation is loaded using the `README` file contained with
the data.

### _Q4.b_

```{r}

# Question 4.b ------------------------------------------------------------

DLcurve.plot(
  tfr_phase2_mcmc,
  country = "Peru",
  nr.curves = 50,
  pi = 95
)

DLcurve.plot(
  tfr_phase2_mcmc,
  country = "Ecuador",
  nr.curves = 50,
  pi = 95
)

```

From the above graphs comparing the Phase II double logistic models for
Peru and Ecuador, we see that the TFR decrements maintains higher values
across TFR in Peru for both the median and 95% PI, suggesting that
fertility is declining faster in Peru than Ecuador We can also observe
this trend by looking directly at the TFR for both countries over time:

```{r}
knitr::kable(
  tfr_peru_ecu,
  booktabs = TRUE,
  digits = 3,
  caption = "TFR over time"
)
```


### _Q4.c_

```{r}

# Question 4c -------------------------------------------------------------

tfr_traj_peru <- get.tfr.trajectories(tfr_pred, country = "Peru")[-(1:2), ]
tfr_traj_ecu <- get.tfr.trajectories(tfr_pred, country = "Ecuador")[-(1:2), ]

prob_tfr_peru_higher <- rowMeans(tfr_traj_peru > tfr_traj_ecu)

prob_tfr_peru_higher_tbl <- tibble(
  period_start = as.integer(names(prob_tfr_peru_higher)) - 3,
  prob_peru_higher = prob_tfr_peru_higher
)

prob_tfr_peru_higher_all <-
  sum(apply(tfr_traj_peru > tfr_traj_ecu, 2, all)) / ncol(tfr_traj_peru)

```

By getting the TFR trajectories for Peru and Ecuador, we can find the
posterior predictive probability for many conditions.

First, we can determine the probability of Peru having a higher TFR than
Ecuador in each five-year period from 2020 through 2095 by finding the mean
number of times Peru has a higher TFR than Ecuador across all simulations:

```{r}
knitr::kable(
  prob_tfr_peru_higher_tbl,
  booktabs = "TRUE",
  digits = 3,
  col.names = c("Period Start", "Pr(peru > edu)")
)
```

We can also find the probability that the TFR of Peru will be higher than that
of Ecuador in all five-year periods from 2020 through 2095 by finding the
mean number of simulations where Peru has a higher TFR than Ecuador for all
periods. This value is calculated to be
**`r round(prob_tfr_peru_higher_all, 3)`**.


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
