---
title: "Homework 04"
author: "Spencer Pease"
date: "4/27/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(ggplot2)
library(tibble)

# Helper functions
scale_integer <- function(x, n = 1) seq(ceiling(x[1]), floor(x[2]), by = n)

# Control randomness
set.seed(9876)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

prior_a <- 1
prior_b <- 1

n_married <- 112
divorced_n_obs <- 43

```

### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

divorced_posterior <- function(n) {
  rbeta(n, prior_a + divorced_n_obs, n_married - divorced_n_obs + prior_b)
}

```

We can generally define the posterior distribution as the product
of a likelihood and prior function:

$$
\begin{aligned}
\text{posterior} &\propto \text{likelihood} \times \text{prior} \\
P(\theta \mid X = x) &\propto P(X =x \mid \theta) \times P(\theta)
\end{aligned}
$$

where, given the parameters of the problem, the likelihood takes the form of a
binomial distribution, and the prior probability follows a uniform distribution
(or, a beta distribution with $\alpha = \beta = 1$):

$$
\begin{aligned}
X \mid \theta &\sim \text{Binomial}(X, \theta) \\
\theta &\sim \text{Beta}(1,1)
\end{aligned}
$$

We can define our binomial as the function $f_{bin}(k, n, p)$, where this
describes getting exactly $k$ successes in $n$ trials, each with a probability
$p$ of occurring.

$$
f_{bin}(k, n, p) =
f_{bin}(k, n, \theta) =
P(X=k \mid \theta; n) =
{n \choose k} \theta^{k}(1-\theta)^{n-k}
$$

Since the prior, $\theta$, is a beta distribution, it acts as a conjugate
prior for the binomial likelihood function. From this, we know that the
posterior distribution will also be a beta distribution in the form:

$$
f_{post}(\theta, k, \alpha, \beta) =
P(\theta \mid X = k; \alpha, \beta) =
\text{Beta}(\alpha + k, n-k+\beta)
$$

From the problem, we know $k = `r divorced_n_obs`$ divorces occurred over the
period 2005-2015 from a sample of $n = `r n_married`$ married people in 2005.
This leads us to our final analytic posterior distribution:

$$
\begin{aligned}
P(\theta \mid X = k; \alpha, \beta, n) &= \text{Beta}(\alpha + k, n-k+\beta) \\
P(\theta \mid X = 43; 1, 1, 112) &= \text{Beta}(1 + 43, 112 - 43 + 1) \\
&= \text{Beta}(44, 70)
\end{aligned}
$$


### _Q1.b_

```{r}

# Question 1b -------------------------------------------------------------

divorced_n_sims <- 1000

## NOTE: Not the way to do this, but still useful reference
#
# divorced_prob_obs <- divorced_n_obs / n_married
# divorced_dist <- function(n, p) rbinom(n, n_married, p)
# divorced_prior_dist <- function(n) rbeta(n, 1, 1)
#
# divorced_prob_draws <- divorced_prior_dist(divorced_n_sims)
# divorced_dist_draws <- divorced_dist(divorced_n_sims, divorced_prob_draws)
#
# divorced_post_dist <- divorced_prob_draws[divorced_dist_draws == divorced_n_obs]

divorced_posterior_draws <- divorced_posterior(divorced_n_sims)

divorced_post_tbl <- tibble(
  Mean = mean(divorced_posterior_draws),
  Median = quantile(divorced_posterior_draws, .5),
  `95% Low` = quantile(divorced_posterior_draws, .025),
  `95% High` = quantile(divorced_posterior_draws, .975)
)

```

```{r}
knitr::kable(
  divorced_post_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Posterior distribution summary"
)

```



### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

post_draws_density <- function(draws) {

  ggplot(data = NULL, aes(draws)) +
  geom_bar() +
  scale_x_binned(n.breaks = 16) +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Posterior Density Estimate",
    y = "Count"
  )

}

divorced_post_plot <-
  post_draws_density(divorced_posterior_draws) +
  labs(
    subtitle = "Probability of divorce by 2015 given marriage in 2005",
    x = "Probability of Divorce"
  )

divorced_post_plot

```


## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

study_obs <- c(
   2.1, 9.8, 13.9, 11.3, 8.9, 15.7, 16.4, 4.5, 8.9, 11.9, 12.5, 11.1, 11.6,
  14.5, 9.6,  7.4,  3.3, 9.1,  9.4,  6.0, 7.4, 8.5,  1.6, 11.4,  9.7
)

study_sd_true <- 4
study_dist <- function(n, mean) rnorm(n, mean = mean, sd = study_sd_true)
study_prior_dist <- function(n) rnorm(n, mean = 10, sd = 3)

```

```{r}

study_n_sims <- 1000

study_mean_draws <- study_prior_dist(study_n_sims)
study_dist_draws <- study_dist(study_n_sims, study_mean_draws)

study_post_dist <- study_mean_draws[study_dist_draws == mean(study_obs)]

study_post_tbl <- tibble(
  Mean = quantile(study_post_dist, .5),
  `95% Low` = quantile(study_post_dist, .025),
  `95% High` = quantile(study_post_dist, .975)
)


```

```{r}
knitr::kable(
  study_post_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Posterior distribution summary"
)

```


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
