---
title: "Homework 06"
author: "Ihsan Kahveci"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: yes
    number_sections: yes
  word_document: default
---

\tableofcontents 
\newpage

```{r setup, include=FALSE}
gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.asp = gr, fig.width = 12)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(tidyverse)
library(bayesLife)
library(bayesPop)
library(migest)

# Data folders
e0.dir <- "../data/e0/sim03092016"
tfr.dir <- "../data/tfr/sim01192018"
pop.dir <- "../data/pop/sim05222022"
mig.file <- "../data/WPP2019_Period_Indicators_Medium.csv"


# Control randomness
set.seed(57)
options(scipen = 999)
```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

# Run this once to get pop predictions
# pop.pred <- pop.predict(
#                end.year = 2100, start.year = 1950, present.year = 2020,
#                wpp.year = 2019, output.dir = pop.dir, nr.traj = 50,
#                inputs = list(tfr.sim.dir = tfr.dir,
#                              e0F.sim.dir = e0.dir,
#                              e0M.sim.dir = "joint_"),
#                keep.vital.events = FALSE, replace.output=TRUE)

pop_sim_pred <- get.pop.prediction(pop.dir)

```

### _Q1.a_

In `bayes.pop`, One can specify the quantities derived from population projections using `pop.expressions`.

> As another example, the potential support ratio can be defined as "PFR[5:13] / PFR[14:27]". (Ševčíková, Raftery 2016)

Ševčíková, H., & Raftery, A. E. (2016). bayesPop: Probabilistic Population Projections. Journal of Statistical Software, 75(5), 1–29. https://doi.org/10.18637/jss.v075.i05


```{r}

# Question 1a -------------------------------------------------------------

# Defined using `?pop.expressions`
over65_exp <- "PAU[14:27]"
support_exp <- "PAU[5:13] / PAU[14:27]"

```

Using our converged life expectancy at birth and total fertility rate simulations, we can create probabilistic projections of the following population quantities:

_Note that the "Potential Support Ratio" is defined as:
$\frac{\text{people aged 20-64}}{\text{people aged 65 and over}}$_


```{r}
pop.trajectories.plot(
  pop_sim_pred, "Australia",
  sex = "both",
  sum.over.ages = TRUE,
  main = "Australia Total Population"
)
```

### _Q1.b_

```{r}
# Question 1b -------------------------------------------------------------

pop.trajectories.plot(
  pop_sim_pred, "Australia",
  sex = "male",
  sum.over.ages = TRUE,
  main = "Australia Total Male Population"
)

```

### _Q1.c_

```{r}
# Question 1c -------------------------------------------------------------

pop.trajectories.plot(
  pop_sim_pred, "Australia",
  expression = over65_exp,
  sex = "both",
  sum.over.ages = TRUE,
  main = "Australia Total Population over 65"
)

```

### _Q1.d_

```{r}
# Question 1d -------------------------------------------------------------

pop.trajectories.plot(
  pop_sim_pred, "Australia",
  expression = support_exp,
  sex = "both",
  sum.over.ages = TRUE,
  main = "Australia Potential Support Ratio",
  show.legend = FALSE)

legend("topright", c("median",  "80% PI", "90% PI", "observed"), 
       col = c("red", "red", "red", "black"), 
       lty = c(1,2,3, 1), cex = 1, bty = "n", lwd = c(2,2,2,2))

```

Looking at the projections of potential support ratio, we see continued
declines into the future, though eventually at a slower rate. This indicates
Australia's age structure is likely to continue shifting to older ages. From a social security perspective, this means that Australia's number of workers per retiree is going to decrease from 2020-2100. 

## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

au_mig <- read_csv(mig.file) %>%
  filter(Location == "Australia") %>%
  transmute(year = MidPeriod - 3, nmr = CNMR, nm = NetMigrations) %>%
  filter(year < 2020)

```

### _Q2.a_

```{r}

# Question 2a -------------------------------------------------------------
au_mig %>%
  select(-nm) %>% 
  knitr::kable(
  booktabs = TRUE, digits = 3,
  col.names = c("Period start", "CNMR"),
  caption = "Crude net migration rate (CNMR) for Australia, 1950-2100"
)

```

### _Q2.b_


```{r}
# Question 2b -------------------------------------------------------------
au_ts <- au_mig %>%
  pull(nmr) %>%
  ts()

au_model = arima(au_ts, order=c(1,0,0))
```

I fit an order $1$ autoregressive model to the subset of Australia Crude Net Migration Rates
obtained from UN, and extract some model parameters below.
_Note that the AR(1) model was fit using the "mle" method._

```{r, echo=FALSE}
modelsummary::msummary(au_model, 
                       estimate = "{estimate} [{conf.low}, {conf.high}]", 
                       statistic = NULL, fmt = 2, output = "kableExtra",
                       caption = "AR(1) Model for Australia CNMR, 1950-2020")  %>%
  kableExtra::kable_styling(latex_options = "hold_position")

```



### _Q2.c_

Using these model parameters, we can find an analytic solution for the
predictive probability distribution of net migration rates in Australia for
2020-2025, which takes the form:

$$ (F_{t+1} - \mu) = \rho*(F_{t} - \mu) + \epsilon $$

$$ \sigma^2 =\frac{ \sum^{N}_{i=1} (\hat{y_i} - y_i)^2 }{N} $$


```{r}
# Question 2c -------------------------------------------------------------

au_mig_last = au_mig$nmr[length(au_mig$nmr)]
au_sd = sqrt(au_model$sigma2)
au_mean = au_model$coef["intercept"]
au_ar1 = au_model$coef["ar1"]
au_mean_pred = unname(au_ar1) * (au_mig_last - au_mean) + au_mean
au_pred_dist <- qnorm(seq(.001, .999, .001), mean = au_mean_pred, sd = au_sd)

au_pred_tbl <- tibble(
  Mean = au_mean_pred,
  Median = au_mean_pred,
  `2.5% CI` = Mean - 1.96 * au_sd,
  `97.5% CI` = Mean + 1.96 * au_sd)
```


- $\sigma^2 = `r au_model$sigma2`$  
- $F_{2020} = `r au_mean_pred`$ 

```{r, echo=FALSE}
knitr::kable(
  au_pred_tbl,
  booktabs = TRUE,
  digits = 3,
  caption = "Predictive Distribution of Australia Crude Net Migration Rate, 2020-2025") %>%
  kableExtra::kable_styling(latex_options = "hold_position")

```


### _Q2.d_

```{r}

# Question 2d -------------------------------------------------------------

au_pred_dist <- qnorm(seq(.0005, .9995, .001), mean = au_mean_pred, sd = au_sd)

au_2020_sample <- rnorm(1000, mean = au_mean_pred, sd = au_sd)

au_2020_tbl <-
  tibble(
    Analytic = au_pred_dist,
    Sample = au_2020_sample
  ) %>%
  pivot_longer(everything(), names_to = "dist", values_to = "value")

```

We can also sample from this distribution to show it does indeed follow what
we predict analytically:

```{r}
ggplot(au_2020_tbl, aes(x = value, fill = dist)) +
  geom_histogram(
    data = filter(au_2020_tbl, dist == "Sample"),
    aes(y = ..density..),
    binwidth = .5,
    alpha = .5) +
  geom_density(
    data = filter(au_2020_tbl, dist == "Analytic"),
    alpha = .2) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predictive Distributions of Crude Net Migration Rate (Analytic and Sample)",
    subtitle = "Australia, 2020-2025",
    x = "Crude Net Migration Rate",
    y = "Density",
    fill = "Type"
  )

```


### _Q2.e_

Assuming that age and sex distribution of migrants has not changed from 2012, which is the last
time the UN reported age/sex-specific net migration rates. 

```{r}
# Question 2e -------------------------------------------------------------

library(wpp2012)
data("migrationF", package = "wpp2012")
data("migrationM", package =  "wpp2012")

au_mig12 <- bind_cols( 
  migrationF %>%
  filter(country == "Australia") %>%
  select(age, migF = `2015-2020`), 
  migrationM %>%
  filter(country == "Australia") %>%
  select(migM = `2015-2020`)) %>% 
  mutate(migTotal = migF + migM, 
         pasmr = migTotal/sum(migTotal))


au_mig20 = tibble(age = au_mig12$age, pasmr = au_mig12$pasmr) %>% 
  mutate(asmr = pasmr * au_mean_pred)

```

Based on age distribution of net migration stocks in 2012, I decomposed the projected migration rate for 2020 by age groups using the formula:

$$ \text{ASMR}_{20} = \text{PASMR}_{12} * \text{CNMR}_{20} $$

```{r}
au_mig20 %>%
  knitr::kable(
  booktabs = TRUE, digits = 2,
  col.names = c("Age Grop", "PASMR", "ASMR"),
  caption = "Predicted Age-spesific Migration Rates (ASMR) for Australia, 2020-2025", 
)

```


```{r}
au_mig20 %>%
  mutate(age = seq(0, 100, 5)) %>%
  ggplot(aes(age, asmr)) + 
  geom_line(color="coral", size=3) +
  theme_bw(base_size = 15) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Predicted Age-spesific Migration Rates (ASMR)",
    subtitle = "Australia, 2020-2025",
    x = "Age-spesific Migration Rate",
    y = "Age" ) +
  scale_x_continuous(n.breaks = 10)
```


\newpage

### _Q2e_ Alternative 

```{r}
rc_decomposition = function(x, params){
  a1 = params[["a1"]]
  alpha1 = params[["alpha1"]]
  a2 = params[["a2"]]
  alpha2 = params[["alpha2"]]
  mu = params[["mu2"]]
  lambda = params[["lambda2"]]
  c = params[["c"]]
  
  mx = a1 * exp(-alpha1 * x) + #first component
    a2 * exp(-alpha2 * (x - mu) - exp(-lambda * (x - mu))) + #second component
    c #constant component
  
  return(mx)
} 

rc_params = deframe(migest::rc_model_fund)
```

Migration schedules without a retirement peak may be represented by a "reduced" model with seven parameters, because in such instances the  _post-labor force_ component of the model is omitted (Rogers & Castro 1981). The simplified 7 parameter model:

$$ M(x) = a_1e^{-\alpha_1x} + a_2e^{-\alpha_2(x-\mu) - e^{-\lambda(x-\mu)}} + c $$ 

`migest` package provides fundamental parameters for Rogers-Castro migration schedule. I plug them into the equation above: 

$$ M(x) = `r rc_params["a1"]` e^{-`r rc_params["alpha1"]`x} + `r rc_params["a2"]`e^{-`r rc_params["alpha2"]`(x-`r rc_params["mu2"]`) - e^{-`r rc_params["lambda2"]`x-`r rc_params["mu2"]`)}} + `r rc_params["c"]` $$

The result is modeled age schedule for migration rates. Hence, if we multiply the resulting vector with the predicted migration rate for 2020-2025, we get age-spesific migration rates for 2020-2025. We need to standardize the output so that it sums to 1. 

```{r}
au_mig20_alt = tibble(age = 0:104) %>%
  mutate(pasmr = rc_decomposition(age, rc_params),
         pasmr = pasmr/sum(pasmr), 
         asmr = pasmr * au_mean_pred) 

  au_alt_grouped = au_mig20_alt %>%
  mutate(age_group = rep(1:21, each=5))%>%
  group_by(age_group) %>%
  summarise(asmr = mean(asmr)) %>%
  mutate(age_group = factor(age_group, labels = au_mig12$age))

```




```{r}
au_alt_grouped %>%
  knitr::kable(
  booktabs = TRUE, digits = 2,
  col.names = c("Age Grop", "ASMR"),
  caption = "Rogers-Castro Model Age-spesific Migration Rates (ASMR) for Australia, 2020-2025", 
)
```

```{r}
au_mig20_alt %>%
  ggplot(aes(x=age, y=asmr )) +
  geom_line(color = "coral", size = 3) +
  theme_bw(base_size = 15) + 
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Rogers-Castro Model Age-spesific Migration Rates (ASMR)",
    subtitle = "Australia, 2020-2025",
    x = "Age-spesific Migration Rate",
    y = "Age" ) +
  scale_x_continuous(n.breaks = 10) 
```



\newpage



### _Q2.f_

For this question, I created 50 different migration rate estimates for 2025 based on the AR(1) model _Q2.c_. Then, using the same age schedule from _Q2.e_, I created age-spesific migration rates for each sample. Here is the pseudo-equation: 

$$ \text{ASMR}_{2025} = \text{TotalPop}_{2020} \times \text{NMR}_{2025} \times \text{PASMR}_{2012}$$
$$\text{NMR}_{20205} = N(\mu = `r au_mean_pred`, \sigma = `r au_sd`)  $$

```{r}

# Question 2f -------------------------------------------------------------
data("pop", package="wpp2019")
au_pop20 = pop %>% filter(name == "Australia") %>% pull(`2020`)
mig_est = rnorm(50,  mean = au_mean_pred, au_sd)
# storing age categories as a sepearate variable
age_cat = au_mig12$age
pasmr = au_mig12 %>% select(age, pasmr)

au_mig25 = 
  tibble(sample = 1:50,
       pop20 = au_pop20,
       nmr25 = mig_est) %>%
  mutate(mig25 = au_pop20 * mig_est) %>%
  slice(rep(1:n(), each = 21)) %>%
  mutate(age = rep_len(age_cat, length.out = n()), .after=sample) %>% 
  left_join(pasmr, by = "age") %>%
  mutate(mig_counts25 = mig25*pasmr)


```


```{r fig.asp=1.5, fig.width=7.5}
ggplot(au_mig25, aes(x = mig_counts25)) +
  geom_histogram() +
  facet_wrap(vars(age), ncol = 3) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Projected Age-Specific Net Migration Numbers",
    subtitle = "Australia, 2020-2025",
    x = "Net number of migrants",
    y = "Density"
  )
```

\newpage 

### _Q2.g_

We will use the same code we used to generate population projections in _Q1_, except we will plug  our migration estimates into the `pop.predict` function. `pop.predict` function has a built-in function to disagregated total net migration into age-spesific migration by applying a Rogers-Castro schedule. Hence, we only need to plug the total migration estimates for 2025. 

The challenge in this question is to transform our predicted trajectories into a format that `pop.predict` function understands. Here is the relevant section from the vignette:

> migMtraj, migFtraj, migtraj
Comma-delimited CSV file with male/female age-specific migration trajectories, or total migration trajectories (migtraj). If present, it replaces deterministic projections given by the mig* items. It has a similar format as e.g. e0M.file with columns “LocID”, “Year”, “Trajectory”, “Age” (except for migtraj) and “Migration”. For a five-year simulation, the “Age” column must have values “0-4”, “5-9”, “10-14”, ..., “95-99”, “100+”. In an annual simulation, age is given by a single number between 0 and 100.



Australia Country Code: 36

```{r}
# Question 2g --------------------------------------------------------------
out.dir = "../data/au_2025"
mig.traj = "../data/mig_trajectories.csv"

mig_out  = 
  au_mig25 %>% 
  mutate(LocID = "36", Period = "2020-2025", Year = "2023") %>%
  #mutate(mig25 = mig25*1000) %>%
  select(LocID, Year, Trajectory=sample, Migration=mig25) %>%
  distinct(Trajectory, .keep_all = TRUE)


#write_csv(mig_out,mig.traj)



# # Run this once to get pop predictions
# pop.pred <- pop.predict(
#                end.year = 2100, start.year = 1950, present.year = 2020,
#                wpp.year = 2019, output.dir = out.dir, nr.traj = 50,
#                countries = "Australia",
#                inputs = list(tfr.sim.dir = tfr.dir,
#                              e0F.sim.dir = e0.dir,
#                              e0M.sim.dir = "joint_",
#                              migtraj = mig.traj),
#                keep.vital.events = FALSE, replace.output=FALSE)

au_pred_25 <- get.pop.prediction(out.dir)



outM = bayesPop::pop.byage.table(au_pred_25, country = "Australia", 
                                sex=c("male"), year = 2025)
outF =  bayesPop::pop.byage.table(au_pred_25, country = "Australia", 
                                sex=c("female"), year = 2025)



tblM = outM %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "age") %>%
  select(-`0.1`, -`0.9`) 


tblF = outF %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "age") %>%
  select(-`0.1`, -`0.9`)

```

\newpage 


The following projections include estimated migration counts for 2025. This is for the male population by age:

```{r}
tblM %>%
  knitr::kable(
  booktabs = TRUE, digits = 2,
  col.names = c("Age", "Median", "2.5% PI", "97.5 PI"),
  caption = "Predicted MalePopulation for Australia, 2020-2025")
```



This is for the female population by age:

```{r}
tblF %>%  
  knitr::kable(
  booktabs = TRUE, digits = 2,
  col.names = c("Age", "Median", "2.5% PI", "97.5 PI"),
  caption = "Predicted Female Population for Australia, 2020-2025")


```


\newpage 

# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
