---
title: "Homework 02"
author: "Ihsan Kahveci"
date: "2020-04-18"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(tidyverse, quietly = TRUE)
library(wpp2019)
library(LifeTables)
library(demogR)
library(MortalityLaws)
library(Metrics)

# helper functions
## matrix printing for latex
write_matex <- function(x, digits = 3) {
  # From: https://stackoverflow.com/a/54088015/8866058
  x <- round(x, digits = digits)
  mat_string <- apply(x, 1, function(y) paste(y, collapse = "&"))
  paste("\\begin{bmatrix}", paste0(mat_string, collapse = "\\\\"), "\\end{bmatrix}")
}

## shortcut for matrix multiplication and exponents
"%^%" <- function(A, n) {
  if (n == 1) {
    A
  } else {
    A %*% (A %^% (n - 1))
  }
}

## creating a Leslie matrix from fertilty and survival rates
make_leslie_matrix <- function(f, s) {
  if (length(f) != length(s)) {
    stop("f and s must be the same length")
  }

  n_size <- length(f)
  l_mat <- matrix(0, nrow = n_size, ncol = n_size)

  l_mat[1, ] <- f
  diag(l_mat[-1, ]) <- s[1:(n_size - 1)]
  l_mat[n_size, n_size] <- s[n_size]

  l_mat
}

## plotting log mortality models
plot_log_mort <- function(data, ...) {
  ggplot(data, aes(...)) +
    geom_point() +
    geom_line() +
    theme_bw() +
    theme(
      text = element_text(family = "serif", size = 15),
      legend.position = "bottom"
    ) +
    labs(
      title = "Age-Specific Log Mortality Rates",
      subtitle = "Peru, Females, 2015-2020",
      x = "Age (years)",
      y = "Log Mortality Rate",
      color = "Model"
    )
}
```


## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

pop_table <- tibble(
  age = c("1", "2", "3+"),
  pop = c(18, 17, 14) * 1000,
  fr = c(0, .9, .2),
  surv = c(.65, .75, .15)
)

knitr::kable(
  pop_table,
  booktabs = TRUE,
  caption = "Q1 one-sex closed population",
  col.names = c(
    "Age",
    "Population ($N_x$)",
    "Fertility Rate ($\\tilde{F}_x$)",
    "Survival Prob. ($s_x$)"
  ),
  eval = FALSE
)

```


### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

CBR <- pop_table %>%
  mutate(births = pop * fr,
         person_years = pop * surv) %>%
  summarise(cbr = sum(births) / sum(person_years)) %>%
  pull(cbr)

```

The crude birth rate ($CBR$) is defined as the number of births over the
person-years lived in the period $[T_1, T_2]$. Since our period is a single
year, we can calculate $CBR$ as:

$$
CBR = \sum\frac{N_x \tilde{F}_x}{N_xs_x}
$$

where we sum over all age groups. The crude birth rate for this population in
the next time period is then **`r round(CBR, 3)`**.


### _Q1.b_

```{r}
# Question 1b -------------------------------------------------------------
## function for convertion fertility rate to expected number of female births
f_tilde_2_asfr <- function(F_tilde, srb, Sxm1, Nxm1, Nx, q0) {
  F_tilde * (1 + srb) * 2/(1 + Sxm1 * (Nxm1/Nx)) / (1 - q0/2)}

pop_asfr_1 <- 0
pop_asfr_2 <- f_tilde_2_asfr(0.9, 1.05, .65, 18000, 17000, 1 - 0.65)
pop_asfr_3 <- f_tilde_2_asfr(0.2, 1.05, .75, 17000, 14000, 1 - 0.65)

pop_asfr <- c(pop_asfr_1, pop_asfr_2, pop_asfr_3)

TFR <- sum(pop_asfr)
tfr_eqn <- paste0(round(pop_asfr, 3), collapse = " + ")

```


The total fertility rate in the population in the period $[T_1, T_2]$ is
defined as the sum of the age-specific fertility rates across all age groups,
multiplied by the length of the age interval, $n$. With $T_2 - T_1 = n = 1$,
the total fertility rate represents the single-year cohort total fertility rate:

$$
\text{TFR}[T_1, T_1 + 1] = \sum {}_{1}F_{x}[T_1, T_1 + 1]
$$

We can convert between $\tilde{F}_x$ and ${}_{1}F_{x}$ using the equation

$$
\tilde{F}_x = {}_{1}F_{x} \times
\frac{1}{1 + SRB} \times
\frac{1}{2}\left( 1 + s_{x-1} \frac{N_{x-1,t}}{N_{x,t}} \right) \times
\left( 1 - \frac{q_0}{2} \right)
$$

where we assume $SRB = 1.05$ and take $q_0 = 1 - s_0$. After converting to
age-specific fertility rates, we calculate a total fertility rate of
$`r tfr_eqn` =$ **`r round(TFR, 3)`** for this population.


### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

pop_leslie <- make_leslie_matrix(pop_table$fr, pop_table$surv)
```

The Leslie matrix, $L$, for this population is defined as:

$$
L =
\begin{bmatrix}
\tilde{F}_{A-3} & \tilde{F}_{A-2} & \tilde{F}_{A-1} \\
s_{A-3}         & 0               & 0 \\
0               & s_{A-2}         & s_{A-1}
\end{bmatrix}
= `r write_matex(pop_leslie)`
$$

Where $(A-1)+$ is the highest age group that can be reached in this population,
$3+$, $s_x$ denotes the probability of survival to the next age group for age
group $x$, and $\tilde{F}_{x}$ is the expected number of female births to a
woman age $x$, who survives to the next time interval.


### _Q1.d_

```{r}

# Question 1d -------------------------------------------------------------

pop_t0 <- matrix(pop_table$pop)
pop_t1 <- pop_leslie %*% pop_t0

```


We can project this population forward using the _cohort-component method of
population projection_, which states that the age-specific populations one time
period ahead ($N_{t+1}$) can be calculated from the matrix multiplication of the
age-specific population in the current period ($N_t$) and the Leslie matrix
($L$) of the population. The population by age one period forward from our
given initial population is then:

$$
\begin{aligned} N_{t+1} &= L N_t \\
&= `r write_matex(pop_leslie)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t1)`
\end{aligned}
$$


### _Q1.e_

```{r}

# Question 1e -------------------------------------------------------------
pop_t2 <- (pop_leslie %^% 2) %*% pop_t0

```

This method can be extended to projecting age-specific population $k$ periods
ahead by raising the Leslie matrix to the $k^{\text{th}}$ power ($L^k$). Our
given population, projected $2$ periods into the future is then:

$$
\begin{aligned} N_{t+2} &= L^{2} N_t \\
&= `r write_matex(pop_leslie)`^{2} `r write_matex(pop_t0)` \\
&= `r write_matex(pop_leslie %^% 2)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t2)`
\end{aligned}
$$


### _Q1.f_

```{r}

# Question 1f -------------------------------------------------------------

CBR_t2 <- pop_table %>%
  mutate(
    pop = as.vector(pop_t1),
    births = pop * fr,
    person_years = pop * surv
  ) %>%
  summarise(cbr = sum(births) / sum(person_years)) %>%
  pull(cbr)

pop_asfr_t2_1 <- 0
pop_asfr_t2_2 <- f_tilde_2_asfr(0.9, 1.05, .65, pop_t1[1], pop_t1[2], 1 - 0.65)
pop_asfr_t2_3 <- f_tilde_2_asfr(0.2, 1.05, .75, pop_t1[2], pop_t1[3], 1 - 0.65)

pop_asfr_t2 <- c(pop_asfr_t2_1, pop_asfr_t2_2, pop_asfr_t2_3)

TFR_t2 <- sum(pop_asfr_t2)
tfr_eqn_t2 <- paste0(round(pop_asfr_t2, 3), collapse = " + ")


```

The crude birth rate for this population from time period 1 to time period 2 is
**`r round(CBR_t2, 3)`**.

The total fertility rate between time periods 1&2 is $`r tfr_eqn_t2` = **`r round(TFR_t2, 3)`** for this population. 

### _Q1.g_

```{r}

# Question 1g -------------------------------------------------------------

pop_right_eigen <- eigen(pop_leslie)
dominant_right_index <- which.max(abs(pop_right_eigen$values))

pop_iroi <- log(pop_right_eigen$values[dominant_right_index])

```

From the theorem that $N_t$ converges to $\lambda^t u$ as $t$ approaches
infinity, $log(\lambda)$ is the _instantaneous rate of increase of the
population_. Here, $\lambda$ is defined as the dominant **right eigenvalue** of the
Leslie matrix, or for the equation:

$$
Lv = \lambda v
$$

it is the eigenvalue $\lambda$ with the largest magnitude. For our calculated
Leslie matrix, the instantaneous rate of increase is **`r round(pop_iroi, 3)`.**

### _Q1.h_

```{r}

# Question 1h -------------------------------------------------------------

pop_sad <- matrix(pop_right_eigen$vectors[, dominant_right_index])

```

Again, from the the formula $\lambda^t u$, $u$ is the _stable age distribution_,
and is defined as the dominant **right eigenvector** of the Leslie matrix, which is
the column vector $v$ from the eigendecomposition of $L$ corresponding to the
eigenvalue $\lambda$ with the largest magnitude.

For our calculated Leslie matrix, the stable age distribution is
$`r write_matex(pop_sad)`$.


### _Q1.i_

```{r}

# Question 1i -------------------------------------------------------------

pop_left_eigen <- eigen(t(pop_leslie))
dominant_left_index <- which.max(abs(pop_left_eigen$values))

pop_repv <- matrix(pop_left_eigen$vectors[, dominant_left_index])

```

The reproductive value vector ($v$) is a vector of expected the number of future
offspring of an individual for each age group. A theorem states that $v$ is
the dominant **left eigenvector** of the Leslie matrix for the population. The left
dominant eigenvector of a matrix $A$ is equivalent to the right dominant
eigenvector of the transpose of matrix, $A^\top$. So, in the formula:

$$ L^{\top}u = \kappa u $$

the dominant eigenvector $u$ represents the reproductive values. For our Leslie
matrix, the reproductive value matrix is then $`r write_matex(pop_repv)`$.


\newpage


## _Q2_
### _Q2.a_

```{r}
# Question 2 ---------------------------------------------------------------
## Question 2a -------------------------------------------------------------
data(mxF)
peru_mx <- mxF %>%
  filter(name == "Peru") %>%
  select(age, mx = `2015-2020`)

# Builds a life table by using the mortality rate 
peru_LT = 
  LifeTables::lt.mx(peru_mx$mx, age = peru_mx$age)$lt %>% 
  as_tibble() %>%
  mutate(age = 
           paste0(Age, "-", Age + 4) %>% 
           magrittr::inset(c(1, 2, length(Age)), c("0", "1-4", "100+")), .before=1) %>%
  mutate(age = as_factor(age))

peru_mx %>% 
  knitr::kable(
  booktabs = TRUE,
  escape = TRUE,
  digits = 3,
  col.names = c("Age", "$_nq_x$"),
  eval = FALSE,
  caption = "Age-specific Mortality Rates of Peru female population, 2015-2020")

```


### _Q2.b_

First, we need to calculate a life table from mortality rates. 

```{r, warning=FALSE}
# Question 2b -------------------------------------------------------------
LT_colnames <- c(
  "Age",
  "$a_x$",
  "$_{5}m_x$",
  "$_{1}q_x$",
  "$_{1}s_x$",
  "$_{5}d_x$",
  "$l_x$",
  "${}_{5}L_x$",
  "${}_{5}T_x$",
  "$e_x$"
)

peru_LT %>% 
  select(-Age)  %>%
  knitr::kable(
    booktabs = TRUE,
    col.names = LT_colnames,
    eval = FALSE,
    digits = c(0, 1, 3, 3, 3, 0, 0, 2, 0, 0, 2),
    caption = "Life table for 2015-2020 Peru female population")


# Create standard 5-year age goup npx and nqx mortality
peru_0_to_5 =peru_LT %>%
  filter(Age < 5) %>%
  summarise(age = "0-4", nqx = 1 - prod(npx), npx = prod(npx))

peru_q0 <- peru_0_to_5$nqx[1]

peru_nqx <- peru_LT %>% 
  filter(Age >= 5) %>%
  select(age, nqx, npx) %>% 
  bind_rows(peru_0_to_5, .)


peru_nqx%>% 
  knitr::kable(
    booktabs = TRUE,
    col.names = c("Age", "$_nq_x$", "$_ns_x$"),
    eval = FALSE,
    digits = c(0, 2, 2),
    caption = "Peru $_ns_x$: 0 and 1-4 combined")
```

Then, we will use, we will use $_ns_x$ and $q_0$ values to convert between $\tilde{F}_x$ and ${}_{1}F_{x}$ using the equation.

```{r, warning=FALSE}
data(popF)
data(sexRatio)
data(tfr)
data(percentASFR)

peru_pop <- popF %>% 
  filter(name == "Peru") %>% 
  select(age, pop = `2015`) %>% 
  mutate(row_id = row_number(), .before=1)

peru_srb <- sexRatio %>% filter(name == "Peru") %>% pull(`2015-2020`)
peru_tfr <- tfr %>% filter(name == "Peru") %>% pull(`2015-2020`)

asfr_2_f_tilde <- function(asfr, srb, Sxm1, Nxm1, Nx, q0) {
  asfr * (1 / (1 + srb)) * .5 * (1 + Sxm1 * (Nxm1 / Nx)) * (1 - q0 / 2)}

peru_Fx <- percentASFR %>%
  filter(name == "Peru") %>%
  select(age=age, pasfr = `2015-2020`) %>%
  mutate(asfr = pasfr * peru_tfr / 5) %>%
  right_join(peru_pop, by = "age", ) %>%
  replace_na(list(pasfr = 0, asfr = 0)) %>%
  left_join(peru_nqx, by = "age") %>%
  mutate(
    f_tilde =
      asfr_2_f_tilde(asfr, peru_srb, lag(npx), lag(pop), pop, peru_q0)
  ) %>%
  replace_na(list(f_tilde = 0)) %>%
  arrange(row_id) %>%
  select(-row_id)

```

To calculate ${}_{5}\tilde{F}_x$, we first use the provided proportional
age-specific fertility rate and total fertility rate for Peru in 2015 to get
age-specific fertility rate with the formula:

$$ \frac{TFR[T_1, T_2] \times {}_{n}PASFR_x}{n} = {}_{n}F_x[T_1, T_2] $$

where $n = 5$ and $[T_1, T_2] = [2015, 2020]$. Then we use the provided
population and previously calculated mortality rates ${}_{5}s_x$ and ${}_{5}q_0$
to calculate ${}_{5}\tilde{F}_x$ using the formula:

$$
{}_{n}\tilde{F}_x = {}_{n}F_x \times
\frac{1}{1 + SRB} \times
\frac{1}{2} \left( 1 + {}_{n}s_x \frac{{}_{5}N_{x-1,t}}{{}_{5}N_{x,t}} \right) \times
\left( 1 - \frac{{}_{n}q_0}{2} \right)
$$

Note that ${}_{5}q_0$ was calculated as $1 - {}_{5}s_0$, which was in turn
calculated from ${}_{1}s_0 \times {}_{4}s_1$.

With these calculations, our resulting ${}_{5}\tilde{F}_x$ is:

```{r}
knitr::kable(
  select(peru_Fx, age, f_tilde, asfr, pop, npx),
  booktabs = TRUE,
  col.names =
    c("Age", "${}_{5}\\tilde{F}_x$", "${}_{5}F_x$", "${}_{5}N_x$", "${}_{5}s_x$"),
  eval = FALSE,
  digits = c(0, 3, 3, 0, 3),
  caption = paste(
    "Expected number of live female births per woman per five-year period",
    "in Peru, 2015-2020"
  )
)
```


### _Q2.c_

```{r}
peru_leslie <- with(peru_Fx, make_leslie_matrix(f_tilde, npx))

peru_leslie_idx <- which(peru_leslie != 0, arr.ind = TRUE)

peru_leslie_tbl <-
  peru_leslie_idx %>%
  as_tibble() %>%
  mutate(value = peru_leslie[peru_leslie_idx]) %>%
  arrange(row, col)
```

Using our calculated ${}_{5}\tilde{F}_x$ and ${}_{5}s_x$, we can build a Leslie
matrix for this population:

```{r}
knitr::kable(
  peru_leslie_tbl,
  booktabs = TRUE,
  col.names = c("Row", "Column", "Value"),
  digits = c(0, 0, 3),
  caption = "Leslie matrix for Peru females, 2015-2020, (sparse format)"
)
```

### _Q2.d_

```{r}
# Question 2d -------------------------------------------------------------
peru_pop_t1 <- peru_leslie %*% matrix(peru_Fx$pop)

```

Using the given female population in Peru, 2015-2020 (_table 3_), we can
calculate the population one 5-year period ahead to 2020-2025, assuming fertility and
mortality are constant over time, and that there is no migration.

$$ N_{t+1} = LN_{t} = `r write_matex(matrix(peru_pop_t1), 0)` $$


### _Q2.e_

```{r}
# Question 2.e -------------------------------------------------------------
peru_pop_t2 <- (peru_leslie %^% 2) %*% matrix(peru_Fx$pop)

```

Under the same assumptions as the previous question, we can instead project the
population 10 years (2 time periods) into the future.

$$ N_{t_2} = L^2N_{t} = `r write_matex(matrix(peru_pop_t2), 0)` $$

## _Q3_

### _Q3.a_

```{r}
# Question 3 -------------------------------------------------------------
## Net migration is per 1000 person-years
data(migration)

peru_net_mig <- migration %>%
  filter(name == "Peru") %>%
  pull(`2015-2020`)

peru_mig <- peru_pop %>%
  mutate(
    mig_rate = (peru_net_mig / 1000) * (pop / sum(pop)),
    mig_rate_fmt = formatC(mig_rate, digits = 2, format = "e"),
    mig_num = pop * mig_rate)

```

Under the assumption that the age-specific migration rates follow the same
distribution as the population distribution, we can calculate age-specific
migration rates using the provided net migration rate (after converting the rate
from _migrations per 1,000 person-years_ to _migrant per person-year_).

```{r}
peru_mig %>%
  select(age, mig_rate_fmt) %>%
  knitr::kable(
  booktabs = TRUE,
  col.names = c("Age", "${}_{5}G_x$"),
  eval = FALSE,
  caption = "Age-specific migration rate in Peru, 2015-2020")
```


### _Q3.b_

```{r}
# Question 3b -------------------------------------------------------------
peru_pop_mig_t1 <- `+`(
  (peru_leslie %*% matrix(peru_mig$pop)),
  (.5 * (1 + peru_leslie)) %*% matrix(peru_mig$mig_num)
)

peru_pop_mig_t2 <- (peru_leslie %^% 2) %*% matrix(peru_mig$pop) +  (peru_leslie %^% 2) %*% matrix(peru_mig$mig_num)

peru_pop_mig_tbl <- tibble(
  age = peru_pop$age,
  mig_pop_2025 = peru_pop_mig_t1,
  mig_pop_2030 = peru_pop_mig_t2
)


```

To incorporate migration into the population projection to 2030, I used the
assumption that all migration happens half-way through a time interval:

$$
\begin{aligned}
N_{t+1} &= LN_t + L^{\frac{1}{2}}G_t \\
&\approx  LN_t + \frac{1}{2}(I + L)G^t
\end{aligned}
$$
and

$$
\begin{aligned}
N_{t+2} &= L^2N_t + L^{\frac{2}{2}}G_t \\
&\approx  L^2N_t + LG^t
\end{aligned}
$$
where $N$ and $G$ refer to female population and migrants, and $I$ is the
identity matrix. Using this equation, we calculate the with-migration female
population in 2020 and 2030:

```{r}
knitr::kable(
  mutate_if(peru_pop_mig_tbl, is.numeric, round),
  booktabs = TRUE,
  col.names = c("Age", "2025", "2030"),
  caption = "Peru female with-migration population projections"
)
```




### _Q3.c_

```{r}
peru_pop_all_tbl <- peru_pop_mig_tbl %>%
  mutate(
    pop_2025 = peru_pop_t1,
    pop_2030 = peru_pop_t2) %>%
  select(age, pop_2025, mig_pop_2025, pop_2030, mig_pop_2030)

```

We can compare our population projections for 2020 and 2030 with migration to
the same projections without migration:

```{r}
knitr::kable(
  mutate_if(peru_pop_all_tbl, is.numeric, round),
  booktabs = TRUE,
  col.names = c("Age", "2025", "+Mig. 2025", "2030", "+Mig. 2030"),
  caption = paste(
    "Comparison of Peru female population projections,",
    "with and without migration"
  )
)
```

Here we see that the with-migration projections are always larger than the
without-migration projections, since Peru has a net positive migration
rate in all age groups (_Table 7_).



## _Q4_

### _Q4.a_

```{r}
# Question 4a -------------------------------------------------------------
model_gompertz2 <- lm(log(nmx) ~ age, data = filter(peru_LT, Age >= 50))

model_gompertz <- with(
  filter(peru_LT, Age >= 50),
  MortalityLaws::MortalityLaw(x = Age, mx = nmx, law = "gompertz")
)

model_makeham <- with(
  filter(peru_LT, Age >= 50),
  MortalityLaw(x = Age, mx = nmx, law = "makeham")
)

coef_gompertz <- coef(model_gompertz)
coef_makeham <- coef(model_makeham)

```

I used 5-year age groups based on the Life Table extracted in question 2. A log-linear fit applied to each of the models, defined below, along with the estimated model parameters after fitting our
observed data:

**Gompertz**

$(\alpha, \beta) =$ (`r round(coef_gompertz, 3)`)

$$
\begin{aligned}
\mu(x) &= \alpha e^{\beta x} \\
log[\mu(x)] &= log(\alpha) + \beta x
\end{aligned}
$$


**Gompertz-Makeham**

$(\alpha, \beta, \gamma) =$ (`r round(coef_makeham, 4)`)

$$
\begin{aligned}
\mu(x) &= \gamma + \alpha e^{\beta x} \\
log[\mu(x) - \gamma] &= log(\alpha) + \beta x
\end{aligned}
$$

_Note: both of these models were fit using the MortalityLaws package, optimizing
the function $log^2(\frac{est.}{obs.})$._


### _Q4.b_

```{r}

# Question 4b -------------------------------------------------------------

peru_LT_model <- peru_LT %>%
  rename(obs_qx = nqx, obs_mx = nmx) %>%
  mutate(
    Gompertz = predict(model_gompertz, x = Age),
    Makeham = predict(model_makeham, x = Age)
  ) %>%
  select(age, everything())

peru_LT_model %>%
  select(-obs_qx , age, Observed = obs_mx, everything()) %>%
  mutate(age = as_factor(age)) %>%
  pivot_longer(-age, names_to = "model", values_to = "mx") %>%
  filter(model %in% c("Observed", "Gompertz", "Makeham")) %>%
  plot_log_mort(x = age, y = log(mx), color = model, group = model) 

```

Plotting both fitted rates against the observed rates, we see that the
additional constant in the _Gompertz-Makeham_ enables the model to follow the
observed trends in child and young adult mortality much closer. This presents
a good case for using _Gompertz-Makeham_ over just _Gompertz_.


### _Q4.c_

```{r}

# Question 4c -------------------------------------------------------------
n = 
model_HP_mx <- with(peru_LT, MortalityLaw(x = Age, mx = nmx, law = "HP"))
model_HP_qx <- with(peru_LT, MortalityLaw(x = Age, qx = nqx, law = "HP"))

peru_LT_model <- peru_LT_model %>%
  mutate(
    n = lead(Age) - Age,
    `Heligman-Pollard` = predict(model_HP_mx, x = Age),
    `Heligman-Pollard2` = -1 * log(1 - predict(model_HP_qx, x =ge)) / n)

coef_HP <- coef(model_HP_mx)

```

The _Heligman-Pollard_ model uses three terms to capture child mortality,
the adult accident bump, and old-age mortality to predict odds of death:

$$
\frac{q_x}{1 - q_x} = A^{(x+B)^2} + D e^{-E(log(x) - log(F))^2} + GH^x
$$

Fitting this model with our observed mortality rates, we get the model
parameters:

(`r paste0(LETTERS[1:8], collapse = ", ")`) =
`r formatC(coef_HP, format = "g", digits = 3)`

_Note: this model was fit using the MortalityLaws package, optimizing
the function $log^2(\frac{est.}{obs.})$, and using $M_x$ as an input (internal
$M_x$ to $q_x$ conversion)._

```{r}
peru_LT_model %>%
  select(-obs_qx, age, Observed = obs_mx, everything()) %>%
  pivot_longer(-age, names_to = "model", values_to = "mx") %>%
  filter(model %in% c("Observed", "Heligman-Pollard")) %>%
  plot_log_mort(x = age, y = log(mx), color = model, group=model)

```

This model does a great job of capturing the high child mortality and gets
closer to capturing the adult accident hump, following the observed data
closely.


### _Q4.d_

```{r}
# Question 4d -------------------------------------------------------------
coaleDemenyLTW <- demogR::cdmltw(sex = "F")
peru_mx_0to95 <- peru_LT %>% filter(Age < 100) %>% pull(nmx)

best_match_lt <-
  coaleDemenyLTW[["nmx"]] %>%
  apply(1, function(x) rmse(x, log(peru_mx_0to95))) %>%
  which.min()

peru_LT_model <- peru_LT_model %>%
  mutate(
    obs_qx_95 = obs_qx,
    standard_qx = c(coaleDemenyLTW[["nqx"]][best_match_lt, ], NA)
  )

peru_LT_model[["obs_qx_95"]][21:22] <- c(1, NA)

model_brass <- lm(
  qlogis(obs_qx_95) ~ qlogis(standard_qx),
  data = filter(peru_LT_model, Age < 95)
)

peru_LT_model <- peru_LT_model %>%
  mutate(
    brass_pred_fit_qx = c(plogis(model_brass$fitted.values), 1, NA),
    brass_pred_obs_qx = plogis(predict(model_brass, data.frame(standard_qx = obs_qx))),
    brass_pred_fit_mx = -1 * log(1 - brass_pred_fit_qx) / n
  )

coef_brass <- coef(model_brass)

```

The _Brass relational model_ estimates mortality given a "standard" reference
mortality ($q^*(x)$) as an input. This standard mortality comes from the a
_Coale-Demeny West_ model life table, where the life table that was selected
was the one with the lowest RMSE compared to our observed mortality (life
table index #`r best_match_lt`). The model and estimated parameters are shown
below:

$(\alpha, \beta) =$ `r round(coef_brass, 3)`

$$
\text{logit}(q_{\alpha, \beta}(x)) = \alpha + \beta \times \text{logit}(q^*(x))
$$

```{r}
peru_LT_model %>%
  select(
    Age, Observed = obs_qx,
    Brass = brass_pred_fit_qx, Brass2 = brass_pred_obs_qx
  ) %>%
  pivot_longer(-Age, names_to = "model", values_to = "qx") %>%
  filter(model %in% c("Observed", "Brass")) %>%
  plot_log_mort(x = Age, y = log(qx), color = model, group=model) +
  labs(
    title = "Age-Specific Log Probability of Death",
    y = "Log Probability of Death"
  )

```

Plotting the observed and estimated $q_x$ values, we see that the _Brass_ model
also does a good job of capturing high child mortality, but underestimates
the adult accident hump. In the old ages this model loses validity, since it
was fix with a lower terminal age than is present in the observed data.


### _Q4.e_

```{r}
# Question 4.e -------------------------------------------------------------
peru_LT_compare <- peru_LT_model %>%
  mutate(n = lead(Age) - Age) %>%
  select(
    Age, n, 
    Observed = obs_mx,
    Gompertz,
    Makeham,
    `Heligman-Pollard`,
    Brass = brass_pred_fit_mx
  )

# knitr::kable(
#   peru_LT_compare,
#   booktabs = TRUE,
#   digits = 3,
#   caption = "Comparison of models against observed Mx in peru females, 2015-2020"
# )

peru_LT_compare %>%
  pivot_longer(!c(Age, n), names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "Gompertz", "Makeham", "Heligman-Pollard", "Brass")) %>%
  plot_log_mort(x = Age, y = log(Mx), color = model, group=model)

```


```{r}
peru_LT_compare_fit <- peru_LT_compare %>%
  filter(Age < 95) %>%
  select(-Age) %>%
  mutate_all(log) %>%
  summarise_all(~rmse(., Observed)) %>%
  select(-Observed) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "RMSE") %>%
  arrange(RMSE)

knitr::kable(
  peru_LT_compare_fit,
  booktabs = TRUE,
  digits = 3,
  caption = "$RMSE$ of estimated $log(m_x)$ against observed $log(m_x)$"
)
```

Comparing all four of these models to the observed data, both visually and
via the root mean-squared-error of log-transformed $m_x$, shows that the
_Brass_ model most closely follows the observed data.



# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
