---
title: "Homework 02"
author: "Spencer Pease"
date: "4/13/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(wpp2019)

# Helper functions

write_matex <- function(x, digits = 3) {
  # From: https://stackoverflow.com/a/54088015/8866058
  x <- round(x, digits = digits)
  mat_string <- apply(x, 1, function(y) paste(y, collapse = "&"))
  paste("\\begin{bmatrix}", paste0(mat_string, collapse = "\\\\"), "\\end{bmatrix}")
}

"%^%" <- function(A, n) {
  if (n == 1) {
    A
  } else {
    A %*% (A %^% (n - 1))
  }
}

make_leslie_matrix <- function(f, s) {

  if (length(f) != length(s)) {
    stop("f and s must be the same length")
  }

  n_size <- length(f)
  l_mat <- matrix(0, nrow = n_size, ncol = n_size)

  l_mat[1, ] <- f
  diag(l_mat[-1, ]) <- s[1:(n_size - 1)]
  l_mat[n_size, n_size] <- s[n_size]

  l_mat
}

```


## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

pop_table <- tibble(
  age = c("1", "2", "3+"),
  pop = c(18, 17, 14) * 1000,
  fr = c(0, .9, .2),
  surv = c(.65, .75, .15)
)

knitr::kable(
  pop_table,
  booktabs = TRUE,
  caption = "Q1 one-sex closed population",
  col.names = c(
    "Age",
    "Population ($N_x$)",
    "Fertility Rate ($\\tilde{F}_x$)",
    "Survival Prob. ($s_x$)"
  ),
  eval = FALSE
)

```


### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

CBR <- pop_table %>%
  mutate(
    births = pop * fr,
    person_years = pop * surv
  ) %>%
  summarise(cbr = sum(births) / sum(person_years))

```

The crude birth rate ($CBR$) is defined as the number of births over the
person-years lived in the period $[T_1, T_2]$. Since our period is a single
year, we can calculate $CBR$ as:

$$
CBR = \sum\frac{N_x \tilde{F}_x}{N_xs_x}
$$

where we sum over all age groups. The crude birth rate for this population in
the next time period is then **`r round(CBR[[1,1]], 3)`**.


### _Q1.b_

```{r}

# Question 2b -------------------------------------------------------------

f_tilde_2_asfr <- function(F_tilde, srb, Sxm1, Nxm1, Nx, q0) {
  F_tilde * (1 + srb) * 2/(1 + Sxm1 * (Nxm1/Nx)) / (1 - q0/2)
}

pop_asfr_1 <- 0
pop_asfr_2 <- f_tilde_2_asfr(.9, 1.05, .65, 18000, 17000, 1-.65)
pop_asfr_3 <- f_tilde_2_asfr(.2, 1.05, .75, 17000, 14000, 1-.65)

pop_asfr <- c(pop_asfr_1, pop_asfr_2, pop_asfr_3)

TFR <- sum(pop_asfr)
tfr_eqn <- paste0(round(pop_asfr, 3), collapse = " + ")

```


The total fertility rate in the population in the period $[T_1, T_2]$ is
defined as the sum of the age-specific fertility rates across all age groups,
multiplied by the length of the age interval, $n$. With $T_2 - T_1 = n = 1$,
the total fertility rate represents the single-year cohort total fertility rate:

$$
\text{TFR}[T_1, T_1 + 1] = \sum {}_{1}F_{x}[T_1, T_1 + 1]
$$

We can convert between $\tilde{F}_x$ and ${}_{1}F_{x}$ using the equation

$$
\tilde{F}_x = {}_{1}F_{x} \times
\frac{1}{1 + SRB} \times
\frac{1}{2}\left( 1 + s_{x-1} \frac{N_{x-1,t}}{N_{x,t}} \right) \times
\left( 1 - \frac{q_0}{2} \right)
$$

where we assume $SRB = 1.05$ and take $q_0 = 1 - s_0$. After converting to
age-specific fertility rates, we calculate a total fertility rate of
$`r tfr_eqn` =$ **`r round(TFR, 3)`** for this population.


### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

pop_leslie <- make_leslie_matrix(pop_table$fr, pop_table$surv)

```

The Leslie matrix, $L$, for this population is defined as:

$$
L =
\begin{bmatrix}
\tilde{F}_{A-3} & \tilde{F}_{A-2} & \tilde{F}_{A-1} \\
s_{A-3}         & 0               & 0 \\
0               & s_{A-2}         & s_{A-1}
\end{bmatrix}
= `r write_matex(pop_leslie)`
$$

Where $(A-1)+$ is the highest age group that can be reached in this population,
$3+$, $s_x$ denotes the probability of survival to the next age group for age
group $x$, and $\tilde{F}_{x}$ is the expected number of female births to a
woman age $x$, who survives to the next time interval.


### _Q1.d_

```{r}

# Question 1d -------------------------------------------------------------

pop_t0 <- matrix(pop_table$pop)
pop_t1 <- pop_leslie %*% pop_t0

```


We can project this population forward using the _cohort-component method of
population projection_, which states that the age-specific populations one time
period ahead ($N_{t+1}$) can be calculated from the matrix multiplication of the
age-specific population in the current period ($N_t$) and the Leslie matrix
($L$) of the population. The population by age one period forward from our
given initial population is then:

$$
\begin{aligned} N_{t+1} &= L N_t \\
&= `r write_matex(pop_leslie)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t1)`
\end{aligned}
$$


### _Q1.e_

```{r}

# Question 1e -------------------------------------------------------------

pop_t10 <- (pop_leslie %^% 10) %*% pop_t0

```

This method can be extended to projecting age-specific population $k$ periods
ahead by raising the Leslie matrix to the $k^{\text{th}}$ power ($L^k$). Our
given population, projected $10$ periods into the future is then:

$$
\begin{aligned} N_{t+10} &= L^{10} N_t \\
&= `r write_matex(pop_leslie)`^{10} `r write_matex(pop_t0)` \\
&= `r write_matex(pop_leslie %^% 10)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t10)`
\end{aligned}
$$


### _Q1.f_

```{r}

# Question 1f -------------------------------------------------------------

CBR_t10 <- pop_table %>%
  mutate(
    pop = as.vector(pop_t10),
    births = pop * fr,
    person_years = pop * surv
  ) %>%
  summarise(cbr = sum(births) / sum(person_years))

pop_asfr_1_t10 <- 0
pop_asfr_2_t10 <- f_tilde_2_asfr(.9, 1.05, .65, pop_t10[1], pop_t10[2], 1-.65)
pop_asfr_3_t10 <- f_tilde_2_asfr(.2, 1.05, .75, pop_t10[2], pop_t10[3], 1-.65)

pop_asfr_t10 <- c(pop_asfr_1_t10, pop_asfr_2_t10, pop_asfr_3_t10)

TFR_t10 <- sum(pop_asfr_t10)

```

The crude birth rate for this population 10 time periods ahead is
**`r round(CBR_t10[[1,1]], 3)`**.

The total fertility rate 10 time periods ahead, after converting $\tilde{F}_x$
to ${}_{1}F_{x}$, is **`r round(TFR_t10, 3)`** for this population.


### _Q1.g_

```{r}

# Question 1g -------------------------------------------------------------

pop_right_eigen <- eigen(pop_leslie)
dominant_right_index <- which.max(abs(pop_right_eigen$values))

pop_iroi <- log(pop_right_eigen$values[dominant_right_index])

```

From the theorem that $N_t$ converges to $\lambda^t u$ as $t$ approaches
infinity, $log(\lambda)$ is the _instantaneous rate of increase of the
population_. Here, $\lambda$ is defined as the dominant right eigenvalue of the
Leslie matrix, or for the equation:

$$
Lv = \lambda v
$$

it is the eigenvalue $\lambda$ with the largest magnitude. For our calculated
Leslie matrix, the instantaneous rate of increase is **`r round(pop_iroi, 3)`.**

### _Q1.h_

```{r}

# Question 1h -------------------------------------------------------------

pop_sad <- matrix(pop_right_eigen$vectors[, dominant_right_index])

```

Again, from the the formula $\lambda^t u$, $u$ is the _stable age distribution_,
and is defined as the dominant right eigenvector of the Leslie matrix, which is
the column vector $v$ from the eigendecomposition of $L$ corresponding to the
eigenvalue $\lambda$ with the largest magnitude.

For our calculated Leslie matrix, the stable age distribution is
$`r write_matex(pop_sad)`$.


### _Q1.i_

```{r}

# Question 1i -------------------------------------------------------------

pop_left_eigen <- eigen(t(pop_leslie))
dominant_left_index <- which.max(abs(pop_left_eigen$values))

pop_repv <- matrix(pop_left_eigen$vectors[, dominant_left_index])

```

The reproductive value vector ($v$) is a vector of expected the number of future
offspring of an individual for each age group. A theorem states that $v$ is
the dominant left eigenvector of the Leslie matrix for the population. The left
dominant eigenvector of a matrix $A$ is equivalent to the right dominant
eigenvector of the transpose of matrix, $A^\top$. So, in the formula:

$$ L^{\top}u = \kappa u $$

the dominant eigenvector $u$ represents the repoductive values. For our Leslie
matrix, the reproductive value matrix is then $`r write_matex(pop_repv)`$.


## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

data(mxF)

thailand_mx <- mxF %>%
  filter(name == "Thailand") %>%
  select(age, mx = `2015-2020`)

```


### _Q2.a_

```{r,}

# Question 2a -------------------------------------------------------------

plot_mx <- function(data, x, y) {
  ggplot(data, aes(x = {{ x }}, y = {{ y }})) +
    geom_point() +
    geom_line() +
    theme_bw() +
    theme(text = element_text(family = "serif")) +
    labs(
      title = "Age-Specific Mortality Rates",
      subtitle = "Thailand, Females, 2015-2020",
      caption = "Source: WPP2019",
      x = "Age (years)",
      y = "Mortality Rate"
    )
}

thailand_mx_plot <- plot_mx(thailand_mx, age, mx)
thailand_log_mx_plot <- plot_mx(thailand_mx, age, log(mx)) +
  labs(title = "Age-Specific Log Mortality Rates", y = "Log Mortality Rate")

plot(thailand_mx_plot)
plot(thailand_log_mx_plot)

```

One unusual feature of these mortality rates is the the bump in log-mortality
around the 15-20 year group. This is a common feature in male populations
(called the accident hump), but is not as common in female populations. Another
feature to note is the incredibly high infant mortality rate, which is at the
same level as older adults.

### _Q2.b_

```{r}

# Question 2b -------------------------------------------------------------

thailand_l0 <- 100000
lx = numeric(nrow(thailand_mx))
lx[1] <- thailand_l0
n <- c(1, 4, rep(5, 20))

for (i in 2:length(lx)) {
  lx[i] <- (1 - (1 - exp(-1 * n[i - 1] * thailand_mx$mx[i - 1]))) * lx[i - 1]
}

thailand_lt <- thailand_mx %>%
  mutate(
    age_str =
      paste0(age, "-", age + 4) %>%
      inset(c(1, 2, length(age)), c("0", "1-4", "100+")),
    n = n,
    qx = 1 - exp(-1 * n * mx),
    sx = 1 - qx,
    lx = lx,
    dx = lx - if_else(!is.na(lead(lx)), lead(lx), 0),
    ax = n / 2
  ) %>%
  select(age, age_str, n, everything())

thailand_lt[["ax"]][1] <- 0.07 + 1.7 * thailand_lt[["mx"]][1]
thailand_lt[["qx"]][nrow(thailand_lt)] <- 1
thailand_lt[["sx"]][nrow(thailand_lt)] <- 0

thailand_lt <- thailand_lt %>%
  mutate(
    Lx = 5 * if_else(!is.na(lead(lx)), lead(lx), 0) + ax * dx,
    Tx = rev(cumsum(rev(Lx))),
    ex = Tx / lx
  )

```

To derive a life table from the given ${}_{n}M_x$ values, we use the following
formulas:

$$
\begin{aligned}
{}_{n}q_x &= 1 - e^{-n\cdot {}_{n}M_x} \\
{}_{n}s_x &= 1 - {}_{n}q_x \\
l_{x+n}   &= (1 - {}_{n}q_{x}) \cdot l_x \\
{}_{n}d_x &= l_x - l_{x+n} \\
{}_{n}L_x &= l_{x+n} + a_x \cdot {}_{n}d_x \\
T_x       &= \sum_{age=x}^\infty {}_{n}L_x \\
e_x       &= T_x \div L_x
\end{aligned}
$$

Where we are using the approximation ${}_{n}q_x = 1 - e^{-n\cdot {}_{n}M_x}$ to
to get ${}_{n}q_x$ from ${}_{n}M_x$, and $a_x$ is caclulated as $\frac{n}{2}$
except for the 0-1 age group, where it is $0.07 + 1.7 {}_{1}M_0$.

```{r}
thailand_lt_names <- c(
  "Age",
  "$n$",
  "$_{5}m_x$",
  "$_{1}q_x$",
  "$_{1}s_x$",
  "$l_x$",
  "$_{5}d_x$",
  "$a_x$",
  "${}_{5}L_x$",
  "${}_{5}T_x$",
  "$e_x$"
)

knitr::kable(
  select(thailand_lt, -age),
  booktabs = TRUE,
  col.names = thailand_lt_names,
  eval = FALSE,
  digits = c(0, 1, 3, 3, 3, 0, 0, 2, 0, 0, 2),
  caption = "Life table for 2015-2020 Thailand female population"
)
```


### _Q2.c_

The life expectancy at birth is **`r round(thailand_lt$ex[1], 3)`**, and
the life expectancy at age 10 is **`r round(thailand_lt$ex[4], 3)`**

## _Q3_

```{r}

# Question 3 --------------------------------------------------------------

data(tfr)
data(percentASFR)
data(sexRatio)
data(popF)
data(migration)

```


### _Q3.a_

```{r, warning=FALSE}

# Question 3a -------------------------------------------------------------

thailand_pop <- popF %>% filter(name == "Thailand") %>% select(age, pop = `2015`)
thailand_srb <- sexRatio %>% filter(name == "Thailand") %>% pull(`2015-2020`)
thailand_tfr <- tfr %>% filter(name == "Thailand") %>% pull(`2015-2020`)

thailand_fertilty <- percentASFR %>%
  filter(name == "Thailand") %>%
  select(age, pasfr = `2015-2020`) %>%
  mutate(asfr = pasfr * thailand_tfr / 5) %>%
  right_join(thailand_pop, by = "age") %>%
  replace_na(list(pasfr = 0, asfr = 0))

```



### _Q3.b_
### _Q3.c_
### _Q3.d_

### _Q3.e_

```{r}

# Question 3e -------------------------------------------------------------

# Net migration is per 1000 person-years
thailand_net_mig <- migration %>%
  filter(name == "Thailand") %>%
  pull(`2015-2020`)

thailand_mig <- thailand_pop %>%
  mutate(migrants = (thailand_net_mig / 1000) * pop)

```


### _Q3.f_
### _Q3.g_


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
