---
title: "Homework 02"
author: "Spencer Pease"
date: "4/13/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(wpp2019)

# Helper functions

write_matex <- function(x, digits = 3) {
  # From: https://stackoverflow.com/a/54088015/8866058
  x <- round(x, digits = digits)
  mat_string <- apply(x, 1, function(y) paste(y, collapse = "&"))
  paste("\\begin{bmatrix}", paste0(mat_string, collapse = "\\\\"), "\\end{bmatrix}")
}

"%^%" <- function(A, n) {
  # From: https://r.789695.n4.nabble.com/A-function-for-raising-a-matrix-to-a-power-td824525.html
  if (n == 1) {
    A
  } else {
    A %*% (A %^% (n - 1))
  }
}

```


## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

pop_table <- tibble(
  age = c("1", "2", "3+"),
  pop = c(18, 17, 14) * 1000,
  fr = c(0, .9, .2),
  surv = c(.65, .75, .15)
)

pop_table_cols <- c(
  "Age",
  "Population ($N_x$)",
  "Fertility Rate ($\\tilde{F}_x$)",
  "Survival Prob. ($s_x$)"
)

knitr::kable(
  pop_table,
  booktabs = TRUE,
  caption = "Q1 one-sex closed population",
  col.names = pop_table_cols,
  eval = FALSE
)

```


### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

births <- pop_table %>%
  mutate(
    births = pop * fr * surv,
    person_years = pop * surv
  ) %>%
  pull(births) %>% sum()

```


### _Q1.b_

```{r}

# Question 2b -------------------------------------------------------------

TFR <- sum(pop_table$fr)
tfr_eqn <- paste0(pop_table$fr, collapse = " + ")

```


The total fertility rate in the population in the period $[T_1, T_2]$ is
defined as the sum of the age-specific fertility rates across all age groups,
multiplied by the length of the age interval, $n$. With $T_2 - T_1 = n = 1$,
the total fertility rate represents the single-year cohort total fertility rate:

$$
\text{TFR}[T_1, T_1 + 1] = \sum {}_{1}F_{x}[T_1, T_1 + 1]
$$

Taking each age-specific fertility rate, we calculate a total fertility rate of
$`r tfr_eqn` =$ **`r TFR`** for this population.


### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

pop_leslie <- matrix(
  c(pop_table$fr, c(pop_table$surv[1], 0, 0), c(0, pop_table$surv[2:3])),
  nrow = 3,
  ncol = 3,
  byrow = TRUE
)

```

The Leslie matrix, $L$, for this population is defined as:

$$
L =
\begin{bmatrix}
\tilde{F}_{A-3} & \tilde{F}_{A-2} & \tilde{F}_{A-1} \\
s_{A-3}         & 0               & 0 \\
0               & s_{A-2}         & s_{A-1}
\end{bmatrix}
= `r write_matex(pop_leslie)`
$$

Where $(A-1)+$ is the highest age group that can be reached in this population,
$3+$, $s_x$ denotes the probability of survival to the next age group for age
group $x$, and $\tilde{F}_{x}$ is the expected number of female births to a
woman age $x$, who survives to the next time interval.


### _Q1.d_

```{r}

# Question 1d -------------------------------------------------------------

pop_t0 <- matrix(pop_table$pop)
pop_t1 <- pop_leslie %*% pop_t0

```


We can project this population forward using the _cohort-component method of
population projection_, which states that the age-specific populations one time
period ahead ($N_{t+1}$) can be calculated from the matrix multiplication of the
age-specific population in the current period ($N_t$) and the Leslie matrix
($L$) of the population. The population by age one period forward from our
given initial population is then:

$$
\begin{aligned} N_{t+1} &= L N_t \\
&= `r write_matex(pop_leslie)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t1)`
\end{aligned}
$$


### _Q1.e_

```{r}

# Question 1e -------------------------------------------------------------

pop_t10 <- (pop_leslie %^% 10) %*% pop_t0

```

This method can be extended to projecting age-specific population $k$ periods
ahead by raising the Leslie matrix to the $k^{\text{th}}$ power ($L^k$). Our
given population, projected $10$ periods into the future is then:

$$
\begin{aligned} N_{t+10} &= L^{10} N_t \\
&= `r write_matex(pop_leslie)`^{10} `r write_matex(pop_t0)` \\
&= `r write_matex(pop_leslie %^% 10)` `r write_matex(pop_t0)` \\
&= `r write_matex(pop_t10)`
\end{aligned}
$$


### _Q1.f_

### _Q1.g_

```{r}

# Question 1g -------------------------------------------------------------

pop_right_eigen <- eigen(pop_leslie)
dominant_index <- which.max(abs(pop_right_eigen$values))

pop_iroi <- log(pop_right_eigen$values[dominant_index])

```

From the theorem that $N_t$ converges to $\lambda^t u$ as $t$ approaches
infinity, $log(\lambda)$ is the _instantaneous rate of increase of the
population_. Here, $\lambda$ is defined as the dominant right eigenvalue of the
Leslie matrix, or for the equation:

$$
Lv = \lambda v
$$

it is the eigenvalue $\lambda$ with the largest magnitude. For our calculated
Leslie matrix, the instantaneous rate of increase is **`r round(pop_iroi, 3)`.**

### _Q1.h_

```{r}

# Question 1h -------------------------------------------------------------

pop_sad <- matrix(pop_right_eigen$vectors[, dominant_index])

```

Again, from the the formula $\lambda^t u$, $u$ is the _stable age distribution_,
and is defined as the dominant right eigenvector of the Leslie matrix, which is
the column vector $v$ from the eigendecomposition of $L$ corresponding to the
eigenvalue $\lambda$ with the largest magnitude.

For our calculated Leslie matrix, the stable age distribution is
$`r write_matex(pop_sad)`$.


### _Q1.i_

```{r}

# Question 1i -------------------------------------------------------------

pop_repv <- t(pop_sad)

```

The reproductive value vector ($v$) is a vector of expected the number of future
offspring of an individual for each age group. A theorem states that $v$ is
the dominant left eigenvector of the Leslie matrix for the population. A left
eigendecomposition for the Leslie matrix is defined as:

$$ uL = \kappa u $$
Taking the transpose of this equation yields:

$$ L^\top u^\top = \kappa u^\top $$

which is similar to the right eigendecomposition, except using the transpose of
the eigenvector. Therefore, the reproductive value vector is the transpose of the
dominant right eigenvector of the Leslie matrix, or the transpose of the stable
age distribution: $`r write_matex2(pop_repv)`$.

## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

data(mxF)
thailand_mx <- mxF %>%
  filter(name == "Thailand") %>%
  select(age, Raw = `2015-2020`) %>%
  mutate(`Log-Transformed` = log(Raw)) %>%
  pivot_longer(-age, names_to = "scale", values_to = "mx")

```


### _Q2.a_

```{r, fig.dim=c(7, 6)}

# Question 2a -------------------------------------------------------------

thailand_mx_plot <- ggplot(thailand_mx, aes(x = age, y = mx)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(scale), nrow = 2, scales = "free_y") +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Age-Specific Mortality Rates",
    subtitle = "Thailand, Females, 2015-2020",
    caption = "Source: WPP2019",
    x = "Age (years)",
    y = "Mortality Rate (mx and log(mx))"
  )

thailand_mx_plot

```

One unusual feature is...

### _Q2.b_
### _Q2.c_

## _Q3_
### _Q3.a_
### _Q3.b_
### _Q3.c_
### _Q3.d_
### _Q3.e_
### _Q3.f_
### _Q3.g_


# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
