---
title: "Homework 03"
author: "Spencer Pease"
date: "4/20/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(MortalityLaws)
library(demogR)
library(MortCast)
library(demography)

# Helper functions
rmse <- function(x, y) sqrt(mean((x - y)^2))
mx_to_qx <- function(mx, n) 1 - exp(-1 * n * mx)
qx_to_mx <- function(qx, n) -1 * log(1 - qx) / n

# Load data
data(mxF, package = "wpp2019")
tha_mort <- mxF %>%
  filter(name == "Thailand") %>%
  select(-country_code, -name) %>%
  pivot_longer(-age, names_to = "period", values_to = "Mx") %>%
  extract(period, "year", regex = "(^[:digit:]{4})", convert = TRUE) %>%
  filter(year < 2020) %>%
  select(year, age, everything()) %>%
  arrange(year, age)

```


## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

tha_mort_2015 <- tha_mort %>%
  filter(year == 2015) %>%
  select(-year) %>%
  mutate(n = lead(age, default = Inf) - age) %>%
  mutate(qx = 1 - exp(-1 * n * Mx))

```

### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

plot_log_mort <- function(data, ...) {

  ggplot(data, aes(...)) +
    geom_point() +
    geom_line() +
    theme_bw() +
    theme(
      text = element_text(family = "serif"),
      legend.position = "bottom"
    ) +
    labs(
      title = "Age-Specific Log Mortality Rates",
      subtitle = "Thailand, Females, 2015-2020",
      x = "Age (years)",
      y = "Log Mortality Rate",
      color = "Model"
    )

}

```

Using the _WPP2019_ package, we can extract the age-specific mortality rates
for females in Thailand in the period 2015-2020. This data is presented in a
table and a log-transformed graph below.

```{r}
knitr::kable(
  select(tha_mort_2015, age, n, Mx),
  booktabs = TRUE,
  col.names = c("Age", "n", "${}_{n}M_x$"),
  caption = "Thailand 2015-2020 female mortality rates",
  eval = FALSE,
  digits = 3
)

```

```{r}
plot_log_mort(tha_mort_2015, x = age, y = log(Mx))
```


### _Q1.b_

```{r}

# Question 1b -------------------------------------------------------------

model_gompertz2 <- lm(log(Mx) ~ age, data = filter(tha_mort_2015, age >= 50))

model_gompertz <- with(
  filter(tha_mort_2015, age >= 50),
  MortalityLaw(x = age, mx = Mx, law = "gompertz")
)

model_makeham <- with(
  filter(tha_mort_2015, age >= 50),
  MortalityLaw(x = age, mx = Mx, law = "makeham")
)

coef_gompertz <- coef(model_gompertz)
coef_makeham <- coef(model_makeham)

```

Two ways of modeling mortality rates are with the _Gompertz_ model and
_Gompertz-Makeham_ model. These models estimate force of mortality $\mu(x)$
using a simple exponential, with the _Gompertz-Makeham_ model including an
additional constant term. A log-linear fit applied to each of the models,
defined below, along with the estimated model parameters after fitting our
observed data:

**Gompertz**

$(\alpha, \beta) =$ (`r round(coef_gompertz, 3)`)

$$
\begin{aligned}
\mu(x) &= \alpha e^{\beta x} \\
log[\mu(x)] &= log(\alpha) + \beta x
\end{aligned}
$$


**Gompertz-Makeham**

$(\alpha, \beta, \gamma) =$ (`r round(coef_makeham, 4)`)

$$
\begin{aligned}
\mu(x) &= \gamma + \alpha e^{\beta x} \\
log[\mu(x) - \gamma] &= log(\alpha) + \beta x
\end{aligned}
$$

_Note: both of these models were fit using the MortalityLaws package, optimizing
the function $log^2(\frac{est.}{obs.})$._


### _Q1.c_

```{r}

# Question 1c -------------------------------------------------------------

tha_mort_2015_model <- tha_mort_2015 %>%
  rename(obs_qx = qx, obs_Mx = Mx) %>%
  mutate(
    Gompertz = predict(model_gompertz, x = age),
    Makeham = predict(model_makeham, x = age)
  ) %>%
  select(age, n, everything())

tha_mort_2015_model %>%
  select(-obs_qx, -n, age, Observed = obs_Mx, everything()) %>%
  pivot_longer(-age, names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "Gompertz", "Makeham")) %>%
  plot_log_mort(x = age, y = log(Mx), color = model)

```

Plotting both fitted rates against the observed rates, we see that the
additional constant in the _Gompertz-Makeham_ enables the model to follow the
observed trends in child and young adult mortality much closer. This presents
a good case for using _Gompertz-Makeham_ over just _Gompertz_.


### _Q1.d_

```{r}

# Question 1d -------------------------------------------------------------

model_HP_mx <- with(tha_mort_2015, MortalityLaw(x = age, mx = Mx, law = "HP"))
model_HP_qx <- with(tha_mort_2015, MortalityLaw(x = age, qx = qx, law = "HP"))

tha_mort_2015_model <- tha_mort_2015_model %>%
  mutate(
    `Heligman-Pollard` = predict(model_HP_mx, x = age),
    `Heligman-Pollard2` = -1 * log(1 - predict(model_HP_qx, x = age)) / n
  )

coef_HP <- coef(model_HP_mx)

```

The _Heligman-Pollard_ model uses three terms to capture child mortality,
the adult accident bump, and old-age mortality to predict odds of death:

$$
\frac{q_x}{1 - q_x} = A^{(x+B)^2} + D e^{-E(log(x) - log(F))^2} + GH^x
$$

Fitting this model with our observed mortality rates, we get the model
parameters:

(`r paste0(LETTERS[1:8], collapse = ", ")`) =
`r formatC(coef_HP, format = "g", digits = 3)`

_Note: this model was fit using the MortalityLaws package, optimizing
the function $log^2(\frac{est.}{obs.})$, and using $M_x$ as an input (internal
$M_x$ to $q_x$ conversion)._

```{r}
tha_mort_2015_model %>%
  select(-obs_qx, -n, age, Observed = obs_Mx, everything()) %>%
  pivot_longer(-age, names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "Heligman-Pollard")) %>%
  plot_log_mort(x = age, y = log(Mx), color = model)

```

This model does a great job of capturing the high child mortality and gets
closer to capturing the adult accident hump, following the observed data
closely.


### _Q1.e_

```{r}

# Question 1e -------------------------------------------------------------

coaleDemenyLTW <- demogR::cdmltw(sex = "F")
tha_Mx_0to95 <- tha_mort_2015 %>% filter(age < 100) %>% pull(Mx)

best_match_lt <-
  coaleDemenyLTW[["nmx"]] %>%
  apply(1, function(x) rmse(x, log(tha_Mx_0to95))) %>%
  which.min()

tha_mort_2015_model <- tha_mort_2015_model %>%
  mutate(
    obs_qx_95 = obs_qx,
    standard_qx = c(coaleDemenyLTW[["nqx"]][best_match_lt, ], NA)
  )

tha_mort_2015_model[["obs_qx_95"]][21:22] <- c(1, NA)

model_brass <- lm(
  qlogis(obs_qx_95) ~ qlogis(standard_qx),
  data = filter(tha_mort_2015_model, age < 95)
)

tha_mort_2015_model <- tha_mort_2015_model %>%
  mutate(
    brass_pred_fit_qx = c(plogis(model_brass$fitted.values), 1, NA),
    brass_pred_obs_qx = plogis(predict(model_brass, data.frame(standard_qx = obs_qx))),
    brass_pred_fit_mx = -1 * log(1 - brass_pred_fit_qx) / n
  )

coef_brass <- coef(model_brass)

```

The _Brass relational model_ estimates mortality given a "standard" reference
mortality ($q^*(x)$) as an input. This standard mortality comes from the a
_Coale-Demeny West_ model life table, where the life table that was selected
was the one with the lowest RMSE compared to our observed mortality (life
table index #`r best_match_lt`). The model and estimated parameters are shown
below:

$(\alpha, \beta) =$ `r round(coef_brass, 3)`

$$
\text{logit}(q_{\alpha, \beta}(x)) = \alpha + \beta \times \text{logit}(q^*(x))
$$

```{r}
tha_mort_2015_model %>%
  select(
    -n, age, Observed = obs_qx,
    Brass = brass_pred_fit_qx, Brass2 = brass_pred_obs_qx
  ) %>%
  pivot_longer(-age, names_to = "model", values_to = "qx") %>%
  filter(model %in% c("Observed", "Brass")) %>%
  plot_log_mort(x = age, y = log(qx), color = model) +
  labs(
    title = "Age-Specific Log Probability of Death",
    y = "Log Probability of Death"
  )

```

Plotting the observed and estimated $q_x$ values, we see that the _Brass_ model
also does a good job of capturing high child mortality, but underestimates
the adult accident hump. In the old ages this model loses validity, since it
was fix with a lower terminal age than is present in the observed data.


### _Q1.f_

```{r}

# Question 1f -------------------------------------------------------------

tha_mort_2015_compare <- tha_mort_2015_model %>%
  select(
    age, n,
    Observed = obs_Mx,
    Gompertz,
    Makeham,
    `Heligman-Pollard`,
    Brass = brass_pred_fit_mx
  )

# knitr::kable(
#   tha_mort_2015_compare,
#   booktabs = TRUE,
#   digits = 3,
#   caption = "Comparison of models against observed Mx in Thailand females, 2015-2020"
# )

tha_mort_2015_compare %>%
  pivot_longer(!c(age, n), names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "Gompertz", "Makeham", "Heligman-Pollard", "Brass")) %>%
  plot_log_mort(x = age, y = log(Mx), color = model)

```


```{r}

tha_mort_2015_compare_fit <- tha_mort_2015_compare %>%
  filter(age < 95) %>%
  select(-age, -n) %>%
  mutate_all(log) %>%
  summarise_all(~rmse(., Observed)) %>%
  select(-Observed) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "RMSE") %>%
  arrange(RMSE)

knitr::kable(
  tha_mort_2015_compare_fit,
  booktabs = TRUE,
  digits = 3,
  caption = "$RMSE$ of estimated $log(M_x)$ against observed $log(M_x)$"
)
```

Comparing all four of these models to the observed data, both visually and
via the root mean-squared-error of log-transformed $M_x$, shows that the
_Heligman-Pollard_ model most closely follows the observed data.


## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

tha_mort_wide <- tha_mort %>% pivot_wider(names_from = year, values_from = Mx)
tha_mort_mat <- tha_mort_wide %>% column_to_rownames("age") %>% as.matrix()

```

### _Q2.a_

Using the _WPP2019 package_, we can extract the female age-specific mortality
rates in Thailand in the for each five year period encompassing 1950-2020.

```{r}

# Question 2a -------------------------------------------------------------

knitr::kable(
  tha_mort_wide,
  booktabs = TRUE,
  digits = 3,
  caption = "Thailand female age-specific mortality rates, 1950-2020"
)

```


### _Q2.b_

```{r}

# Question 2b -------------------------------------------------------------

model_LC <- leecarter.estimate(tha_mort_mat)

LC_lsq_mx <- with(model_LC, apply(
  matrix(bx) %*% matrix(kt, nrow = 1), 2, function(x) ax + x
))
dimnames(LC_lsq_mx) <- dimnames(tha_mort_mat)

tha_mort_lsq <- LC_lsq_mx %>%
  as_tibble(rownames = "age") %>%
  pivot_longer(-age, names_to = "year", values_to = "log_Mx") %>%
  mutate(
    year = as.integer(year),
    age = as.integer(age),
    `Least-squares Estimate` = exp(log_Mx),
  ) %>%
  select(-log_Mx)

tha_LC_compare <- tha_mort %>%
  left_join(tha_mort_lsq, by = c("age", "year")) %>%
  rename(Observed = Mx)

```

The _Lee-Carter_ model is defined as:

$$
log(m_{x,t}) = a_x + k_t b_x + \epsilon_{x,t}
$$

One way to fit this model is the apply a set of constraints to $b_x$ and $k_t$.
Under the constraints $\sum b_x = 1$ and $\sum k_t = 0$, we get:

$$
\hat{a}_x = \frac{1}{T} \sum_{t=1}^T log(m_{x,t}) \\
\hat{k}_t = \sum_{x=0}^{A-1} \left[ log(m_{x,t}) - a_x \right]
$$
With these constraints, we can fit a  least squares regression on $\hat{b}_x$
with an intercept of 0. Below are the estimated parameters and mortality rates

_Note: This model was fit using the MortCast package._

```{r}
model_LC_ls_x_tbl <- tibble(
  age = names(model_LC$ax),
  ax = model_LC$ax,
  least_squares_bx = model_LC$bx
)

model_LC_ls_t_tbl <- tibble(
  year = names(model_LC$kt),
  least_squares_kt = model_LC$kt
)

knitr::kable(
  model_LC_ls_x_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Age", "$\\hat{a}_x$", "$\\hat{b}_x$"),
  eval = FALSE,
  caption = "Lee-Carter model parameter estimates (least squares method)"
)

knitr::kable(
  model_LC_ls_t_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Year", "$\\hat{k}_t$"),
  eval = FALSE,
  caption = "Lee-Carter model parameter estimates (least squares method)"
)

```

```{r, fig.dim=c(7, 8.5)}
tha_LC_compare %>%
  pivot_longer(!c(year, age), names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "Least-squares Estimate")) %>%
  plot_log_mort(x = age, y = log(Mx), color = model) +
  facet_wrap(vars(year), ncol = 3) +
  labs(subtitle = "Thailand, Females, 1950-2020")

```


### _Q2.c_

```{r}

# Question 2c -------------------------------------------------------------

ax_hat <- rowMeans(log(tha_mort_mat))
model_LC_svd <- svd(apply(log(tha_mort_mat), 2, function(x) x - ax_hat), 1, 1)

# Get normalized bx and kt
LC_svd_bx <- model_LC_svd$u / sum(model_LC_svd$u)
LC_svd_kt <- t(model_LC_svd$v) * sum(model_LC_svd$u) * model_LC_svd$d[1]

LC_svd_mx <- apply(LC_svd_bx %*% LC_svd_kt, 2, function(x) ax_hat + x)
dimnames(LC_svd_mx) <- dimnames(tha_mort_mat)

tha_mort_svd <- LC_svd_mx %>%
  as_tibble(rownames = "age") %>%
  pivot_longer(-age, names_to = "year", values_to = "log_Mx") %>%
  mutate(
    year = as.integer(year),
    age = as.integer(age),
    `SVD Estimate` = exp(log_Mx),
  ) %>%
  select(-log_Mx)

tha_LC_compare <- tha_LC_compare %>%
  left_join(tha_mort_svd, by = c("age", "year"))

```

A singular value decomposition can also be used to implement the _Lee-Carter_
model, where the matrix $C = \left[ log(m_{x,t}) - a_x \right]$ is used to get
$b_x$ and $k_t$ with the decomposition:

$$ \text{SVD}(C) = U \Lambda V^\top $$

Where:

$$
b_x = U_{x,1} \div \sum_x U_{x,1} \\
k_t = \left[ V^\top \right]_{1,t} \times \sum U_{x,1} \times \Lambda_1
$$

are the normalized estimates. The estimated parameters and fitted values are
below.

```{r}
model_LC_svd_x_tbl <- tibble(
  age = names(model_LC$bx),
  svd_bx = as.vector(LC_svd_bx)
)

model_LC_svd_t_tbl <- tibble(
  year = names(model_LC$kt),
  svd_kt = as.vector(LC_svd_kt),
)

knitr::kable(
  model_LC_svd_x_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Age", "$b_x$"),
  eval = FALSE,
  caption = "Lee-Carter model parameter estimates (SVD method)"
)

knitr::kable(
  model_LC_svd_t_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Year", "$k_t$"),
  eval = FALSE,
  caption = "Lee-Carter model parameter estimates (SVD method)"
)

```


```{r fig.dim=c(7, 8.5)}
tha_LC_compare %>%
  pivot_longer(!c(year, age), names_to = "model", values_to = "Mx") %>%
  filter(model %in% c("Observed", "SVD Estimate")) %>%
  plot_log_mort(x = age, y = log(Mx), color = model) +
  facet_wrap(vars(year), ncol = 3) +
  labs(subtitle = "Thailand, Females, 1950-2020")

```

Comparing the estimated $M_x$ from the normalized SVD to our observed data, we
see that the estimates also track the observed data well. Note that if we
were to use the unnormalized SVD, the estimated $M_x$ would remain mostly
constant over time, matching the middle years the closest (since those years
are closest to the mean over time).


### _Q2.d_

```{r}

# Question 2d -------------------------------------------------------------

LC_compare_bx_tbl <- model_LC_ls_x_tbl %>%
  left_join(model_LC_svd_x_tbl, by = "age") %>%
  select(-ax) %>%
  mutate(abs_diff = abs(least_squares_bx - svd_bx))

LC_compare_kt_tbl <- model_LC_ls_t_tbl %>%
  left_join(model_LC_svd_t_tbl, by = "year") %>%
  mutate(abs_diff = abs(least_squares_kt - svd_kt))

tha_LC_compare_fit <- tha_LC_compare %>%
  group_by(year) %>%
  select(-age) %>%
  mutate_at(vars(-group_cols()), log) %>%
  summarise_all(~rmse(., Observed)) %>%
  select(-Observed) %>%
  ungroup() %>%
  select(-year) %>%
  summarize_all(mean) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "RMSE") %>%
  arrange(RMSE)

knitr::kable(
  LC_compare_bx_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Age", "LS $\\hat{b}_x$", "SVD $\\hat{b}_x$", "diff"),
  eval = FALSE,
  caption = "Lee-Carter model comparison"
)

knitr::kable(
  LC_compare_kt_tbl,
  booktabs = TRUE,
  digits = 3,
  col.names = c("Year", "LS $\\hat{k}_t$", "SVD $\\hat{k}_t$", "diff"),
  eval = FALSE,
  caption = "Lee-Carter model comparison"
)

knitr::kable(
tha_LC_compare_fit,
booktabs = TRUE,
digits = 4,
caption = "$RMSE$ of estimated $log(M_x)$ against observed $log(M_x)$"
)

```

Looking at the previous two plots, and comparing the root mean-squared-error
of the log-transformed fitted $M_x$ against the observed $M_x$, we see that,
while close, the normalized SVD _Lee-Carter_ method performs the best.


### _Q2.e_

```{r}

# Question 2e -------------------------------------------------------------

data(popF, package = "wpp2019")

mex_pop <- popF %>%
  filter(name == "Mexico") %>%
  select(-country_code, -name, -`2020`) %>%
  extract(age, "age", convert = TRUE) %>%
  column_to_rownames("age") %>%
  as.matrix() %>%
  `*`(1000)

mex_mort_tbl <- mxF %>%
  filter(name == "Mexico") %>%
  select(-country_code, -name) %>%
  pivot_longer(-age, names_to = "period", values_to = "Mx") %>%
  extract(period, "year", regex = "(^[:digit:]{4})", convert = TRUE) %>%
  filter(year < 2020)

# Collapse 0 and 1-4 age group Mx
mex_mort_tbl_u5 <- mex_mort_tbl %>%
  filter(age < 5) %>%
  mutate(
    n = age * 3 + 1,
    sx = 1 - mx_to_qx(Mx, n)
  ) %>%
  group_by(year) %>%
  summarise(sx = prod(sx)) %>%
  mutate(
    age = 0,
    Mx = qx_to_mx(1 - sx, 5)
  ) %>%
  select(age, year, Mx)

mex_mort <- mex_mort_tbl %>%
  filter(age >= 5) %>%
  bind_rows(mex_mort_tbl_u5) %>%
  pivot_wider(names_from = year, values_from = Mx) %>%
  arrange(age) %>%
  column_to_rownames("age") %>%
  as.matrix()

demog_data <- demogdata(
  data = mex_mort,
  pop = mex_pop,
  ages = seq(0, 100, 5),
  years = seq(1950, 2015, 5),
  type = "mortality",
  label = "Mexico",
  name = "Female"
)

model_mex_LC <- demography::lca(demog_data)

forecast_mex <- forecast(model_mex_LC, h = 1, level = 95, se = "innovonly")

forecast_mex_kt <- as_tibble(forecast_mex[["kt.f"]])

forecast_mex_mx <- forecast_mex$rate %>%
  as_tibble() %>%
  mutate_all(as.vector) %>%
  mutate(age = rownames(forecast_mex$rate$Female)) %>%
  select(age, Mx = Female, lower_95 = lower, upper_95 = upper)

forecast_mex_mx_75 <- forecast_mex_mx %>% filter(age == 75)

```

The _Lee-Carter_ method can also be used to obtain a probabilistic forecast of
mortality. Here we use the _demography_ _R_ package to forecast mortality for
females in Mexico in the period 2020-2025. Below are the predicted mortality
index $k_t$ and mortality rate for females in the 75-80 age group. The
confidence level is set to $95\%$.


```{r}
knitr::kable(
  forecast_mex_kt,
  booktabs = TRUE,
  digits = 3,
  eval = FALSE,
  col.names = c("Est. $k_t$", "95% Low", "95% High"),
  caption = "Forecast $k_t$"
)

knitr::kable(
  select(forecast_mex_mx_75, -age),
  booktabs = TRUE,
  digits = 3,
  eval = FALSE,
  col.names = c("Est. $M_x$", "95% Low", "95% High"),
  caption = "Forecast mortality for age group 75-80"
)
```



# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
