---
title: "Homework 05"
author: "Ihsan Kahveci"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: yes
    number_sections: true
---

\tableofcontents 
\newpage

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(tidyverse)
library(bayesLife)
library(bayesPop)

# Data
data("e0F", package = "wpp2019")
e0_sim_dir <- "data/e0/sim03092016"
tfr_sim_dir <- "data/tfr/sim01192018"
pop_sim_dir <- "data/pop/sim05172020"
mig_file <- "data/WPP2019_Period_Indicators_Medium.csv"
can_mig_num_file <- "data/statcan_migration.csv"

e0f_all <- e0F %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    cols = -name,
    names_to = "year",
    values_to = "e0",
    names_pattern = "(.*)-"
  ) %>%
  mutate(year = as.integer(year))

# Control randomness
set.seed(57)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

e0f_nica <- e0f_all %>%
  filter(name == "Nicaragua") %>%
  select(-name) %>%
  mutate(gain = lead(e0) - e0)

knitr::kable(
  e0f_nica,
  booktabs = TRUE, digits = 2, eval = FALSE,
  col.names = c("Period Start", "$e_0$", "Gain"),
  caption = "Nicaragua female life expectancy at birth and observed gains, 1950-2020"
)

```

### _Q1.a_


Following lecture slides, we can model the expected gain $(\ell_{t+1} - \ell{t})$ using double-logistic function for a single country, as follows: 

$$
g(\ell \mid \theta) = 
  \frac{k}{1 + exp(- \frac{2log(9)}{\Delta_2}
    (\ell - \Delta_1 - 0.5\Delta_2))}  - 
  \frac{z - k}{1 + exp(- \frac{2log(9)}{\Delta_4}
    (\ell -\Delta_1 - \Delta_2 - \Delta_3 - 0.5\Delta_4))} 
$$


```{r}

# Question 1a -------------------------------------------------------------
# double-logistic function
# parameters: 4 delta, k, z
dl_gain <- function(l, theta) {

  d1 <- theta[["d1"]]
  d2 <- theta[["d2"]]
  d3 <- theta[["d3"]]
  d4 <- theta[["d4"]]
  k <- theta[["k"]]
  z <- theta[["z"]]

  (k       / (1 + exp( (-2*log(9) / d2) * (l - d1 - .5*d2         )))) +
  ((z - k) / (1 + exp( (-2*log(9) / d4) * (l - d1 - d2 -d3 - .5*d4))))

}

# least squares error function
# takes a function finds the LS between fitted and observed values
ls_err <- function(func, data, obs_vals) {
  
  function(params) {
    fit_vals <- func(data, params)
    sum((fit_vals - obs_vals)^2)
  }
}

```

In order to model gains in life expectancy at birth, we use the six parameter
double logistic model defined in class. Double-logistic model is harder to converge with numerical optimization because there can be many different set of variables that result in the same DL curve. To make the model converge, I used the starting parameters ($\Delta_n$, $k$, and $z$.) of the Markov Chain 1.


```{r}
e0_sim_mcmc<- get.e0.mcmc(e0_sim_dir)
mcmc1 = e0_sim_mcmc$mcmc.list$`1`
#e0.DLcurve.plot(e0_sim_mcmc, country = "Nicaragua")
starting_params <- c(mcmc1$Triangle.ini, mcmc1$k.ini, mcmc1$z.ini)
names(starting_params) <- c("d1", "d2", "d3", "d4", "k", "z")

tibble(params = names(starting_params), values = starting_params) %>% 
  knitr::kable(booktabs = TRUE, digits = 3, eval = FALSE,
               caption = "Inital Values for Least-Squares Double-Logistic Model")
```

```{r}
# removing last row because it has no gain value
opt_input <- e0f_nica %>% slice(-n())
loss_func <- ls_err(dl_gain, opt_input$e0, opt_input$gain)

opt_result <- optim(starting_params, loss_func)

opt_params <- opt_result$par
opt_err_var <- opt_result$value

```


With these starting values, we use `optim()` to minimize the
least-squares-error, loss of the observed gains vs the fitted gains from the
double logistic gain model, which give us the set of optimized parameters
$[`r round(opt_params, 3)`]$ with an error variance of `r round(opt_err_var, 3)`.


### _Q1.b_

```{r}

# Question 1b -------------------------------------------------------------

e0f_nica_gains <- e0f_nica %>%
  rename(obs_gain = gain) %>%
  mutate(fit_gain = dl_gain(e0, opt_params))

cor_nica = cor(e0f_nica_gains$obs_gain[1:13], e0f_nica_gains$fit_gain[1:13])
```

We can plot use our optimized parameters to get estimates of $e_0$ gain, and
compare them to the observed gains to get a sense of how well the optimization
performed:

```{r}
e0f_nica_gains %>%
  drop_na() %>% 
  ggplot(aes(x = obs_gain, y = fit_gain)) +
  geom_point() +
  geom_smooth(formula=y~x, method = "lm", color = "cadetblue") +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Fit vs Observed Gains",
    subtitle = "Nicaragua, 1950-2020",
    x = "Observed Gain (years)",
    y = "Fitted Gain (years)"
  )

```

The model fit considerably well to the observed data, there is a strong correlation (`r round(cor_nica, 2)`) between fitted and observed values. 


### _Q1.c_

$$ \ell_{2020} = \ell_{2015} + g(\ell_{2015} \mid \theta) + \epsilon $$
$$ \sigma^2 =\frac{ \sum^{N}_{i=1} (\hat{y_i} - y_i)^2 }{N} = `r opt_err_var` $$ 


```{r}

# Question 1c -------------------------------------------------------------

nica_e0_2020_mean <- e0f_nica_gains %>%
  filter(year == 2015) %>%
  select(e0, fit_gain) %>%
  rowSums()

nica_e0_2020_sd <- sqrt(opt_err_var)

nica_e0_2020_dist <- qnorm(
  seq(.0005, .9995, .001),
  mean = nica_e0_2020_mean,
  sd = nica_e0_2020_sd
)

```

Using our observed life expectancy at birth for 2015-2020 and the gain in the
same period, we can create an analytic predictive distribution of possible life
expectancy at birth for 2020-2025, using the variance $\sigma^2$ from our model:


```{r}
ggplot(enframe(nica_e0_2020_dist), aes(x = value)) +
  geom_density(fill = "coral", alpha = .25) +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Analytic Predictive Distribution of e(0)",
    subtitle = "Nicaragua, 2020-2025",
    x = "e(0) (years)",
    y = "Density"
  )

```

```{r}
nica_e0_2020_tbl <- tibble(
  Mean = nica_e0_2020_mean,
  Median = median(nica_e0_2020_dist),
  `2.5% PI` = Mean - 1.96 * nica_e0_2020_sd,
  `97.5% PI` = Mean + 1.96 * nica_e0_2020_sd
)

knitr::kable(
  nica_e0_2020_tbl,
  booktabs = TRUE, digits = 3, eval = FALSE,
  caption = "Summary of the predictive distribution of 2020-2025 Nicaragua $e_0$"
)

```


\newpage 

## _Q2_

### _Q2.a_

```{r}

# Question 2a -------------------------------------------------------------

e0_sim_mcmc <- get.e0.mcmc(e0_sim_dir)
e0_sim_pred <- get.e0.prediction(e0_sim_dir)

```



Data from a fully converged simulation from a Bayesian model created with
_BayesLife_ are loaded using the functions `get.e0.mcmc()` and
`get.e0.prediction()`. `get.e0.mcmc()` returns an object containing each MCMC
chain from the simulation, and `get.e0.prediction()` returns an object
containing the summary statistics of the posterior trajectories for the life expectancy created using an input set of MCMC chains.

### _Q2.b_

**Double logistic curve fits for Senegal and Ghana:**

```{r}

# Question 2b -------------------------------------------------------------

e0.DLcurve.plot(e0_sim_mcmc, "Senegal",
  predictive.distr = TRUE, pi = c(95), ylim = c(-2, 8))

e0.DLcurve.plot(e0_sim_mcmc, "Ghana",
  predictive.distr = TRUE, pi = c(95), ylim = c(-2, 8))

```

The model fits well to the both countries, as Predictive Intervals include most of the observed points. It seems like it fits Ghana relatively better considering the distribution of observed points. Although the wide predictive intervals prevents to make strong conclusions, the median lines are still insightful. Senegal reaches the peak 5-year gains when life expectancy at birth is approx. 55 years. Whereas Ghana reaches the peak 5-year gains when life expectancy at birth is approx. 45 years. Also, Senegal has a higher expected 5-year gains on average than Ghana until age 80. Then, their trajectory becomes very similar.Also, Senegal also has a wider probability interval in later ages.

\newpage 

### _Q2.c_

```{r}

# Question 2c -------------------------------------------------------------

e0_traj_tbl <- list(Senegal = "Senegal", Ghana = "Ghana") %>%
  purrr::map(~e0.trajectories.table(e0_sim_pred, country = .x)) %>%
  purrr::map_dfr(~as_tibble(.x, rownames = "year"), .id = "country") %>%
  mutate(year = as.integer(year) - 3) %>%
  filter(year < 2015) %>%
  select(country, `Period start` = year, median) %>%
  pivot_wider(names_from = country, values_from = median)

e0_traj_mean_tbl <- e0_traj_tbl %>%
  pivot_longer(-`Period start`, names_to = "Country", values_to = "e0") %>%
  group_by(Country) %>%
  mutate(gain = lead(e0) - e0)
  
  
  # summarise(`Mean Gain` = mean(gain, na.rm = TRUE),
  #            SD = sd(gain, na.rm = TRUE)) %>%
  # mutate(`5% CI` = `Mean Gain` - 1.96 * SD,
  #        `95% CI` = `Mean Gain` + 1.96 * SD) %>%
  # arrange(desc(`Mean Gain`))

```


Using `e0.trajectories.table` function, one can easily tabulate the tabulate the posterior distribution of trajectories of the life expectancy for list of countries. Note that years before 2018 doesn't have any uncertainty, as they are observed not predicted.
The double-logistic plots indicate that Senegal is experiencing faster increases in life expectancy than Ghana. We can confirm this by examining the reported life expectancies for both countries from 1950 to 2015:

```{r}
knitr::kable(
  e0_traj_tbl,
  booktabs = TRUE, digits = 2, eval = FALSE,
  caption = "Female $e_0$ for Senegal and Ghana, 1950-2015"
)
```

Again, it appears Senegal has a faster increase. We can verify this by
calculating the mean gain over time for each country:

```{r}
e0_traj_mean_tbl %>%
  ggplot(aes(x=`Period start`, y=e0, color= Country)) + 
  geom_line(size=1) + 
  theme_bw() +
  ggtitle("Fitted Life Expectancies for Ghana and Senegal")
```

```{r}
e0_traj_mean_tbl %>%
  ggplot(aes(x=`Period start`, y=gain, color= Country)) + 
  geom_line(size=1) + 
  theme_bw() +
  ggtitle("Fitted Gains in Life Expectancy for Ghana and Senegal")
  
```

```{r}
data("e0F", package = "wpp2015")

e0f_2015 <- e0F %>%
  select(-country_code, -last.observed) %>%
  pivot_longer(
    cols = -country,
    names_to = "year",
    values_to = "e0",
    names_pattern = "(.*)-"
  ) %>%
  mutate(year = as.integer(year)) %>%
  filter(country %in% c("Senegal", "Ghana"))

```
```{r}
e0f_2015 %>%
  mutate(`Period start` = year) %>%
  ggplot(aes(x=`Period start`, y=e0, color= country)) + 
  geom_line(size=1) + 
  theme_bw() +
  ggtitle("Observed Life Expectancy in Senegal and Ghana")
```

Because the existing simulations based on WPP2015 data, I extracted empirical $e_0$ values from `wpp2015` package. The above plot shows that fitted and observed values are mostly similiar; and Senegal shows a faster increase emprically too.


\newpage 

### _Q2.d_

```{r}

# Question 2d -------------------------------------------------------------
# tidying the life expectancy projections 
sen = get.e0.trajectories(e0_sim_pred, "Senegal")[-c(1,2),] 

sen_df = as_tibble(sen) %>% 
  mutate(country = "Senegal", year = seq(2020,2095,5), .before=1)

gha = get.e0.trajectories(e0_sim_pred, "Ghana")[-c(1,2),]

gha_df =   as_tibble(gha) %>% 
  mutate(country = "Ghana", year = seq(2020,2095,5), .before=1)


sen_gha  = bind_rows(sen_df, gha_df) %>% 
  pivot_longer(-c(country, year), names_to = "sim", values_to = "e0") 
  
sen_gha_sum =  sen_gha %>% 
  group_by(country, year) %>%
  summarise(median_e0 = median(e0), 
            low = median_e0 - 1.96*sd(e0),
            high = median_e0 + 1.96*sd(e0)) 

sen_gha_probs = sen_gha %>%
  pivot_wider(id_cols = c(sim, year),  names_from = country, values_from = e0) %>%
  mutate(greater = ifelse(Senegal > Ghana, 1, 0)) %>%
  group_by(year) %>%
  summarise(Senegal = median(Senegal), Ghana = median(Ghana), prob = mean(greater))


```

```{r}
probs = c()
for (i in 1:ncol(sen)){
  out = sen[,i] > gha[,i]
  n_true = length(out[out==TRUE])
  prob = ifelse(n_true == nrow(sen), 1, 0)
  probs = c(probs, prob)
}
probs %>% length()
mean(probs)
```


Assuming gains in life expectancy are independent between countries, we can find
the probability of Senegal having either a higher or lower average life
expectancy at birth than Ghana in each future period by calculating the
number of times Senegal has a higher $e_0$ than Ghana across all simulations:

$$ P(S>G) = \frac{\|S>G\|}{N}$$

```{r}
sen_gha_probs %>%
  filter(year == 2020) %>%
#  pivot_wider(id_cols = year, names_from = country, values_from = median_e0) %>%
  knitr::kable(
  booktabs = TRUE, digits = c(0,1,1,3), eval = FALSE,
  col.names = c("Period Start", "Senegal", "Ghana", "$P(S>G)$"),
  caption = "Senegal and Ghana Median $e_0$ Projections in 2020-2025")
```

For all 1000 simulations:   
  - Probability of Senegal has higher $e_0$ than Ghana in 2020: `r sen_gha_probs$prob[1]`  
  - Probability of Senegal has higher $e_0$ in all 16 periods: `r mean(probs)`  
  
\newpage 

The plot below shows the probability densities of 2020 simulations for Ghana and Senegal. It is clear that the probability of Ghana has higher $e_0$ than Senegal is very low. (shaded area)

```{r}
sen_gha %>%
  filter(year == 2020) %>%
  ggplot(aes(x = e0, fill = country)) +
  geom_density(alpha = 0.3) +
  ggtitle("Predictive Distributions for Life Expectancy in 2020", subtitle = "Ghana vs. Senegal") +
  theme_bw() 
```

\newpage 

We can also observe the future projections of Senegal and Ghana by plotting them over the years. Confidence Intervals calculated using the SD of the simulations for each year: 


```{r}
sen_gha_sum %>%
  ggplot(aes(x=year, y=median_e0, color = country,fill=country)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = low , ymax = high), alpha = 0.1, linetype = "dashed" ) +
  ggtitle("Ghana and Senegal, 2020 - 2100", subtitle = "Probabilistic Trajectories of Life Expectancy") +
  labs(x= " ", y = "Median e0") +
  theme_bw()
```


\newpage

# Appendix



```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
