---
title: "Homework 07"
author: "Spencer Pease"
date: "5/25/2020"
output:
  pdf_document:
    latex_engine: xelatex
    highlight: tango
    df_print: kable
    fig_caption: true
---

```{r setup, include=FALSE}

gr <- 2 / (1 + sqrt(5))

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.asp = gr)
options(knitr.kable.NA = '-')

rm(gr)
```

# Questions

```{r prep, include=FALSE}

# Prep work ---------------------------------------------------------------

# Load libraries
library(dplyr)
library(ggplot2)
library(rtweet)
library(tidycensus)
library(tigris)

```

## _Q1_

```{r}

# Question 1 --------------------------------------------------------------

```

### _Q1.a_

```{r}

# Question 1a -------------------------------------------------------------

bbox_florida <- c(-87.586670, 24.256982, -79.735108, 30.983104)

# Collect Tweets
# streamed_tweets_florida <- stream_tweets(
#   q = bbox_florida,
#   timeout = (60 * 60 * 8),
#   parse = FALSE,
#   file_name = "data/rtweet_stream_florida.json",
# )
#
# tweet_tbl <- parse_stream("data/rtweet_stream_florida.json")
# saveRDS(tweet_tbl, "data/tweet_stream_florida_parsed.RDS")

tweet_tbl <- readRDS("data/tweet_stream_florida_parsed.RDS")


```

This report focuses on collecting Twitter data from Florida. The bounding
box used to encompass Florida is $(`r round(bbox_florida, 3)`)$, created using
[bboxfinder.com][http://bboxfinder.com/].

### _Q1.b_

```{r}

# Question 1b -------------------------------------------------------------

num_tweets <- nrow(tweet_tbl)

time_span_hr <- difftime(
  max(tweet_tbl$created_at), min(tweet_tbl$created_at),
  units = "hours"
  ) %>%
  as.numeric() %>%
  round(digits = 1)

```

After streaming tweets from Florida for around `r time_span_hr` hours,
**`r num_tweets`** were collected. We can inspect the distribution of these
collected tweets in a time series plot:

```{r}
ts_plot(tweet_tbl, by = "1 minutes") +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Number of Streamed Tweets over Time",
    subtitle = paste("Location: Florida, Duration:", time_span_hr, "hours"),
    x = "Time (UTC)",
    y = "Tweets (count)"
  )

```

The periods of zero tweets are likely a result of connection issues when
collecting data, and not representative of the true frequency of tweets.


## _Q2_

```{r}

# Question 2 --------------------------------------------------------------

```

### _Q2.a_

```{r}

# Question 2a -------------------------------------------------------------

# Load US Census API key
CENSUS_API_KEY <- readRDS("~/.uscensus_api_key.RDS")
census_api_key(CENSUS_API_KEY)

```

_Got a census key!_

### _Q2.b_

```{r}

# Question 2b -------------------------------------------------------------

fla_counties <-
  get_acs(
    geography = "county",
    variables = c("Total Population" = "B01001_001"),
    year = 2018,
    state = "Florida",
    geometry = FALSE,
    survey = "acs1"
  ) %>%
  tidyr::separate(NAME, c("county", "state"), sep = ", ") %>%
  mutate(county = stringr::str_remove(county, " County")) %>%
  rename(geoid_county = GEOID, population = estimate) %>%
  select(county, state, geoid_county, population)

```


From the _American Community Survey_, we get total population estimates for
all counties within Florida:

```{r}
knitr::kable(
  fla_counties, booktabs = TRUE,
  col.names = c("County", "State", "GEOID", "Pop. estimate"),
  caption = "Florida counties and total population estimated by ACS1 2018"
)

```

### _Q2.c_

```{r}

# Question 2c -------------------------------------------------------------

```

Estimated total populations of Florida counties can also be visualized:

```{r}
ggplot(fla_counties, aes(x = county, y = population / 1000)) +
  geom_col() +
  theme_bw() +
  theme(
    text = element_text(family = "serif"),
    axis.text.x.bottom = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Distribution of Population across Florida",
    subtitle = "ACS1 2018 Population",
    x = "County",
    y = "Total Population (1000s of people)"
  )

```

_Note: I tried to use a map here, but had issues adding the geometry information
to the data._

## _Q3_

```{r}

# Question 3 --------------------------------------------------------------

```

### _Q3.a_

```{r}

# Question 3a -------------------------------------------------------------

tweet_lat_lon_tbl <- tweet_tbl %>%
  rtweet::lat_lng() %>%
  rename(lon = lng) %>%
  select(lat, lon)

```

Latitude and Longitude information can be extracted from the collected tweet
data with `rtweet:lat_lng()`, which uses all geolocation information in a tweet
to get a coordinate pair. This poses an issue if a twitter user has their
location turned off, or manually set to a location different from where they are
when using Twitter. Using all the geolocation information also means looking at
the bounding box of the tweet, which may overlap with the area of interest, but
not truly be in the area of interest.

We can solve the issue of tweet coming from other locations by subsetting our
data, but it is difficult to avoid the issue of capturing tweets where the
location is manually set to be within our region of interest.

### _Q3.b_

```{r}

# Question 3b -------------------------------------------------------------

# Make function that won't fail when finding GEOIDs for all lat/lon pairs
safe_geolocator <- purrr::possibly(tigris::call_geolocator_latlon, NA_character_)

# Takes a while to run, so save results
# tweet_lat_lon_tbl %>%
#   purrr::pmap_chr(safe_geolocator) %>%
#   saveRDS("data/block_geoids.RDS")

block_geoids <- readRDS("data/block_geoids.RDS")

geocode_tweets <- tweet_tbl %>%
  rtweet::lat_lng() %>%
  rename(lon = lng) %>%
  mutate(
    geoid_block = block_geoids,
    geoid_county = substr(geoid_block, 1, 5)
  ) %>%
  left_join(fla_counties, by = "geoid_county") %>%
  filter(state == "Florida")

num_dropped <- num_tweets - nrow(geocode_tweets)
pct_dropped <- (num_dropped / num_tweets) * 100

```

After adding GEOID data to the tweets, **`r num_dropped`** tweets outside of
Florida were dropped (`r round(pct_dropped, 1)`%).

### _Q3.c_

```{r}

# Question 3c -------------------------------------------------------------

grouped_tweets <- geocode_tweets %>%
  group_by(county, population) %>%
  summarise(tweets = n())

```

With our tweets tagged with the appropriate county-level GEOID, we can
investigate how many tweets are associated with each county:

```{r}
knitr::kable(
  grouped_tweets, booktabs = TRUE,
  col.names = c("County", "Pop. Estimate", "Tweet Count"),
  caption = "Number of collected tweets geo-coded to each Florida county"
)
```


### _Q3.d_

```{r}

# Question 3d -------------------------------------------------------------

tweet_model <- lm(tweets ~ population, data = grouped_tweets)

model_tbl <- tweet_model %>%
  broom::tidy() %>%
  left_join(
    confint(tweet_model) %>% as_tibble(rownames = "term"),
    on = "term"
  ) %>%
  select(-statistic)

```

To ascertain the association between the number of tweets originating from a
county and the county's population, we fit the simple linear model

$$ lm(\text{tweets} \sim \text{population}) $$
which yields the parameters:

```{r}
knitr::kable(
  model_tbl, booktabs = TRUE, digits = 4,
  col.names = c("Term", "Estimate", "Std. Error", "P-value", "2.5% CI", "97.5% CI"),
  caption = "Summary of model fitting the association between tweets and population"
)

```

We can also visualize the model:

```{r}
ggplot(grouped_tweets, aes(x = population / 1000, y = tweets)) +
  geom_point(alpha = .7) +
  geom_smooth(method = "lm") +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Number of Tweets vs Population",
    subtitle = "Florida Counties",
    x = "Total Population Estimate (1000s of people)",
    y = "Tweets (count)"
  )
```

From the data and model, we can say that there is a statistically significant
relationship between the population of a county and the number of tweets from
a county. Removing the few extreme values from the data would significantly
change this relationship, however, so maybe there are some underlying factors
this model doesn't capture (such as tourists coming to Miami and tweeting, but
not being counted as part of the population).

# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```
